Concerted changes in the pediatric single-cell intestinal ecosystem before and after anti-TNF blockade

Version 1

Abstract

Abstract

Crohn’s disease is an inflammatory bowel disease (IBD) commonly treated through anti-TNF blockade. However, most patients still relapse and inevitably progress. Comprehensive single-cell RNA-sequencing (scRNA-seq) atlases have largely sampled patients with established treatment-refractory IBD, limiting our understanding of which cell types, subsets, and states at diagnosis anticipate disease severity and response to treatment. Here, through combining clinical, flow cytometry, histology, and scRNA-seq methods, we profile diagnostic human biopsies from the terminal ileum of treatment-naïve pediatric patients with Crohn’s disease (pediCD; n=14), matched repeat biopsies (pediCD-treated; n=8) and from non-inflamed pediatric controls with functional gastrointestinal disorders (FGID; n=13). To resolve and annotate epithelial, stromal, and immune cell states among the 201,883 baseline single-cell transcriptomes, we develop a principled and unbiased tiered clustering approach, ARBOL. Through flow cytometry and scRNA-seq, we observe that treatment-naïve pediCD and FGID have similar broad cell type composition. However, through high-resolution scRNA-seq analysis and microscopy, we identify significant differences in cell subsets and states that arise during pediCD relative to FGID. By closely linking our scRNA-seq analysis with clinical meta-data, we resolve a vector of T cell, innate lymphocyte, myeloid, and epithelial cell states in treatment-naïve pediCD (pediCD-TIME) samples which can distinguish patients along the trajectory of disease severity and anti-TNF response. By using ARBOL with integration, we position repeat on-treatment biopsies from our patients between treatment-naïve pediCD and on-treatment adult CD. We identify that anti-TNF treatment pushes the pediatric cellular ecosystem towards an adult, more treatment-refractory state. Our study jointly leverages a treatment-naïve cohort, high-resolution principled scRNA-seq data analysis, and clinical outcomes to understand which baseline cell states may predict Crohn’s disease trajectory.

Main

Inflammatory bowel diseases (IBDs) arise when homeostatic mechanisms regulating gastrointestinal (GI) tract tissue integrity, nutrient absorption, and protective immunity are replaced by pathogenic inflammation (Baumgart and Sandborn, 2012; Chang, 2020; Corridoni et al., 2020b; Friedrich et al., 2019; Graham and Xavier, 2020; Selin et al., 2021). The initiating triggers are not fully known, but host genetics and the microbiome are being increasingly appreciated to play important, and in some cases causal roles in the IBDs (Chang, 2020; Cohen et al., 2019; Franzosa et al., 2019; Jain et al., 2021; Limon et al., 2019). Crohn’s disease (CD) presents predominantly in the terminal ileum and the proximal colon, though lesions may develop anywhere along the gastrointestinal tract (Baumgart and Sandborn, 2012; Chang, 2020; Kobayashi et al., 2020; Roda et al., 2020). Among the IBDs, pediatric-onset Crohn’s disease (pediCD) is particularly common (25% of all IBD cases, 60-70% of pediatric IBD) and is a debilitating form due to its early presentation, impact on the terminal ileum and proximal colon, and the lack of disease-specific therapies developed for children (Hyams et al., 1991; Ruemmele et al., 2014; Sýkora et al., 2018; Turner et al., 2012; Ye et al., 2020). In contrast to pediCD, there exists a group of pediatric disorders termed functional gastrointestinal disorders (FGIDs), which include GI symptoms that necessitate endoscopy, but lack laboratory markers, endoscopic findings, and histologic evidence associated with inflammation (Black et al., 2020; Hyams et al., 2016; McOmber and Shulman, 2008; Santucci et al., 2020). Thus, while FGID is a distinct disease entity that does not represent completely healthy tissue, it is a critical non-inflamed age-matched control cohort with which to contextualize the inflammation observed in pediCD.

The current standard of care for pediCD (as with adult CD) is tailored to the patient’s disease location, clinical behavior and severity, though use of prednisone and other immunomodulators, as well as monoclonal antibodies including anti-TNFα, are common (Hyams et al., 1991; Ruemmele et al., 2014; Turner et al., 2012). While targeting TNF is shared across many autoimmune and inflammatory conditions, it is not successful in all patients, and many go on to develop anti-TNF-refractory disease. It is of tremendous importance for the field to precisely understand and characterize for which patients anti-TNF therapy is not necessary, in which it may succeed in controlling disease, and which will likely be refractory to treatment. Several ideas based on individual immunogenicity and pharmacokinetics have been proposed to explain TNF-refractory disease, including gender (M>F), low albumin levels, high BMI, and high baseline C-Reactive Protein (CRP) (Atreya et al., 2020; Digby-Bell et al., 2020). However, no single identifiable clinical or biochemical biomarker reliably predicts disease response versus resistance to anti-TNF treatment (Stevens et al., 2018).

The primary cellular lineages sampled from intestinal biopsies of CD patientsrepresent both the epithelium and lamina propria, and include epithelial cells, stromal cells, hematopoietic cells and neuronal processes whose cell bodies are present outside of these regions (Buisine et al., 2001; Leeb et al., 2003; Leonard et al., 1995; Lilja et al., 2000; Müller et al., 1998; Souza et al., 1999; Stappenbeck and McGovern, 2017; Takayama et al., 2010). Alterations to all cellular lineages have been implicated in CD (Furey et al., 2019; Martin et al., 2019). Histologically, CD is characterized by a granulomatous inflammation, and by alterations in almost every leukocyte cell type studied, including an increase of cytotoxic lymphocytes, alterations in γδ T cells, increases in mast cells and their production of TNF, activation and shifts in antibody isotypes towards IgM and IgG from B cells and plasma cells, and cytokine production by macrophages (Catalan-Serra et al., 2017; Lilja et al., 2000; Meijer et al., 1979; Mitsialis et al., 2020; Müller et al., 1998; Sieber et al., 1984; Takayama et al., 2010). In the stromal compartment, there is evidence for enhanced vascularization and increased expression of ICAM-1 and MAdCAM-1 by vascular beds, substantial remodeling of collecting lymphatics, and altered migratory potential of fibroblasts (Leeb et al., 2003; Souza et al., 1999). Epithelial barrier dysfunction has also been noted, including alterations to mucus production, microvilli, and Paneth cell dysfunction (Buisine et al., 2001; Stappenbeck and McGovern, 2017). Previous studies have therefore identified that essentially all cell types may be meaningfully altered during CD, highlighting the important need to comprehensively chart the concerted cellular changes that define CD, its severity, and its expected treatment course. Importantly, which changes are predictive of more severe disease, response to treatment, or treatment resistance in pediCD remain largely unknown.

Single-cell RNA-sequencing (scRNA-seq) is enhancing our ability to comprehensively map and resolve the cell types, subsets, and states present during health and disease. Recent work has generated cellular atlases of adult CD and UC from patients who were already on treatment and without linking to clinical follow-up studies (Corridoni et al., 2020b, 2020a; Drokhlyansky et al., 2019; Elmentaite et al., 2020; Huang et al., 2019; Kinchen et al., 2018; Martin et al., 2019; Parikh et al., 2019; Smillie et al., 2019). The potential impact of scRNA-seq on our understanding of IBD is evidenced by studies of adult UC, which have identified potential functional roles for poorly understood colonic cell subsets, such as the BEST4+ enterocyte, and identified pathological alterations in UC biopsies compared to healthy controls, including an expansion of microfold-like cells, IL13RA2+IL11+ inflammatory fibroblasts, CD4+CD8+IL17A+ T cells and CD8+GZMK+ T cells (Smillie et al., 2019). A single-cell study of on-treatment adult CD patients comparing non-inflamed and inflamed tissue from surgically-resected bowel identified IgG+ plasma cells, inflammatory mononuclear phagocytes, activated T cells and stromal cells, comprising the “GIMATS” pathogenic cellular module (Martin et al., 2019). This module was used to derive a gene signature associated with resistance to anti-TNF therapy in a distinct cohort profiled by bulk RNA-seq. Very recent work has profiled how fetal transcription factors are reactivated in Crohn’s disease epithelium, leading to significant decreases in mature absorptive epithelial cells and increases in secretory/goblet cells (Elmentaite et al., 2020). Together, these and other studies have demonstrated the power of scRNA-seq to nominate individual and collective cell states that are associated with disease, and underscored the unmet need to apply these techniques to untreated disease and associate them with disease severity in order to more specifically identify pathognomonic and prognostic cell states.

Most comprehensive scRNA-seq atlases of inflammatory disease conditions include patients being treated with a variety of agents, and for which the biopsies often reflect a partial treatment-refractory state to combinations of antibiotics, corticosteroids, immunomodulators, and biologics including anti-TNF monoclonal antibodies. A treatment-naïve single-cell atlas in an inflammatory disease condition linking observed baseline cell clusters with disease trajectory and treatment outcomes has yet to be reported. In order to address this unmet need in pediCD, we created the prospective PREDICT study (Clinicaltrials.gov #NCT03369353) to identify, profile, and understand pediatric IBD, as well as uninflamed FGID controls. Here, we present detailed diagnostic data from the first cohort of 27 patients enrolled on PREDICT, including 14 pediCD and 13 FGID patients, together with flow cytometric and scRNA-seq studies of the cellular composition of the terminal ileum (Figure 1). Furthermore, through prospective annotation of clinical metadata and detailed longitudinal follow-up, we stratify our pediCD cohort by clinically-guided therapeutic decisions separating patients treated with anti-TNF mAbs versus those with biopsy-proven pediCD, but for whom clinical symptoms were sufficiently mild that the treating physician did not prescribe anti-TNF agents (this cohort is termed “Not On Anti-TNF” or “NOA”). Importantly, we were also able to separate the cohort of patients treated with anti-TNF agents into sub-cohorts of those who achieved a full response (FR) to this therapy versus those who achieved only a partial response (PR). Because PREDICT enrolled patients prior to their diagnostic endoscopy, we were able to relate these clinical outcomes to the patients’ cell states at diagnosis, and were further able to integrate matched on-treatment biopsies from these same pediCD patients. We contextualize our findings in pediCD relative to a cohort of FGID patients, which provides an age-matched comparator cohort with clinical GI symptoms, but non-inflammatory disease proven by endoscopy and histologic examination.

PREDICT Study Design with Patient Diagnostic Criteria and Histopathology

a. Study overview depicting clinical and cellular measurements from 13 functional gastrointestinal disorder (FGID) patients and 14 pediatric Crohn’s disease (pediCD) patients. Terminal ileum biopsies were isolated at a treatment-naïve diagnostic visit, and pediCD patients were followed up to determine their anti-TNF response and categorized as not on anti-TNF (NOA), Full Response (FR), or Partial Response (PR) (see Methods). Two panels of flow cytometry allowed for relative frequency quantification of 32 cell types and subsets, and 10X 3’ v2 single-cell RNA-sequencing (scRNA-seq) captured 254,911 total cells including 118 FGID and 305 pediCD end clusters through ARBOL (see Methods).

b. Demographic data, weight, height, and BMI for cohort (see Table 1 and Supplemental Figure 1); Significance testing by Mann-Whitney U-test.

c. Clinical inflammatory laboratory values for cohort (see Table 1 and Supplemental Figure 1); Significance testing by Mann-Whitney U-test.

d. Representative histopathology of FGID (top) and pediCD (bottom) at 10x (scale bar = 100um) and 40x (scale bar = 20um) magnification.

e. Representative treatment history and clinical inflammatory parameters used for determination of NOA (p027), FR (p011) and PR (p018) status (see Methods, Table 1, and Supplemental Figure 1; ADA: adalimumab, INF: infliximab; MTX: methotrexate; Pred: prednisone; mSCD: modified specific carbohydrate diet; EEN: exclusive enteral nutrition).

Several analytical approaches have been developed to enable the generation and interrogation of clusters during the curation of single-cell transcriptomic atlases (Hie et al., 2020). One such method, sub-clustering of broad clusters, has proven to be a powerful tool for isolating highly specific axes of variation that are obscured by analyses whose principal axes of variation are broad cell types (Bakken et al., 2021; La Manno et al., 2021; Sikkema et al., 2022; Tasic et al., 2018; Zeisel et al., 2018). However, while sub-clustering analysis is a powerful tool allowing access to the hierarchy of cell states, this method is manually intensive and there is little consensus or standardization in clustering parameters and annotation methods. To address this issue, we designed a principled, modular, automated, iterative sub-clustering routine made possible by application of parameter scanning methods (Rousseeuw, 1987; Shekhar et al., 2016). We developed this tool, ‘ARBOL’ (named after tree in Spanish), of which iterative tiered clustering (ITC) is a key component, in R and python, integrating with Seurat and Scanpy functions, to make it accessible and easily incorporated into common workflows, and have curated a GitHub repository with illustrative vignettes. Here, we use ARBOL to standardize fine-grained cell state discovery by the creation and cultivation of a tree of cell states, followed by the automated generation of cell names to aid in the annotation of end clusters by unique and descriptive genes, inclusive of a method for identifying clusters containing high or low patient diversity, and for pruning of the resultant tree.

In order to respect the distinct clinical entities of FGID and pediCD, and the co-variation structure that can define rare but important cell subsets and states, we generate two fully-annotated cellular atlases for these pediatric GI diseases. These atlases consist of 94,451 cells for FGID and 107,432 for pediCD. We provide key gene-list resources for further studies, and nominate a vector of lymphoid, myeloid and epithelial cell states that anticipate disease severity and treatment outcomes. We term this vector ‘pediCD T cell, innate lymphocyte, myeloid and epithelial’ (pediCD-TIME). This cellular vector correlates strongly with both the clinical presentation of pediCD severity, and with the distinction between anti-TNF full- or partial-response. The significant changes in cell composition we have discovered to be associated with disease severity are increases of proliferating T cells, cytotoxic NK cells, subsets of monocytes/macrophages, and plasmacytoid dendritic cells (pDCs), accompanied by decreases of metabolically-specialized epithelial cell subsets. We successfully validate this vector in two bulk RNA-seq treatment-naïve IBD cohorts. We contextualize our treatment-naïve pediCD atlas by joining it with the FGID atlas through label-retaining (manual clustering; Random Forest; ARBOL) and label-free (Hellinger distance on UMAP) methods. Finally, we integrate our pediCD atlas with follow-up post-treatment biopsies from our cohort, and with the adult on-treatment CD patients of the “GIMATS” scRNA-seq atlas. Taken together, we identify that anti-TNF treatment over time pushes the overall pediCD cellular ecosystem towards the more treatment-refractory disease state found in adult CD.

Results

Study cohort clinical outcomes

The PREDICT study prospectively enrolled treatment-naïve, previously undiagnosed pediatric patients with GI symptoms necessitating diagnostic endoscopy. The current analysis focuses on patients enrolled in the first year of the study, during which time 14 patients with pediCD and 13 patients with FGID were enrolled and had adequate ileal samples for single-cell analysis (Figure 1; Supplemental Figure 1). After their initial diagnosis, patients with pediCD were followed clinically for up to 3 years (with the determination of anti-TNF response made at 2 years). Patients with FGID were followed clinically only as needed. The median time from diagnosis for the pediCD and FGID cohorts to the time of database lock was 32.5 and 31 months, respectively. Of the pediCD and FGID patients analyzed, the median age at diagnosis was 12.5 years and 16 years respectively (p = 0.095; Mann-Whitney U-test), with no significant differences in gender (Figure 1b; Table 1). Patient weight, height, and BMI z-scores were not significantly different between pediCD and FGID (Figure 1b; Table 1); however, in addition to the diagnostic differences on histologic analysis, several key clinical laboratory values, including C-reactive protein (CRP), Erythrocyte Sedimentation Rate (ESR), hemoglobin concentration, albumin concentration, and platelet count were significantly different between pediCD and FGID (Figure 1c,d; Table 1).

Treatment with anti-TNF agents and response to therapy

Patients with pediCD were initially divided into two cohorts. Those with milder disease characteristics (n = 4) as determined by their treating physician, were not treated with anti-TNF therapies, and are noted as ‘NOA’. For patients with more severe disease (n = 10), anti-TNF therapy (with either infliximab or adalimumab, Table 2) was initiated within 90 days of diagnostic endoscopy. All pediCD patients were followed prospectively and categorized as FR (n = 5) or PR (n = 5) to anti-TNF therapy based on the following criteria: FR was defined as clinical symptom control and biochemical response (measuring CRP, ESR, albumin, and complete blood counts (CBC)), and with a weighted Pediatric Crohn’s Disease Activity Index (PCDAI) score of <12.5 on maintenance anti-TNF therapy with no dose adjustments required (Cappello and Morreale, 2016; Hyams et al., 1991; Sandborn, 2014; Turner et al., 2017, 2012). PR to anti-TNF therapy was defined as a lack of full clinical symptom control as determined by the treating physician or lack of full biochemical response, with documented escalation of anti-TNF therapy or addition of other agents (Figure 1e; NB: patients in our cohort were dose-escalated because of clinical symptoms). Medication timelines and clinical laboratory data through 2 years post-diagnosis for all pediCD patients is shown in Supplemental Figure 1. The designation of FR or PR was made at 2 years of follow-up for all pediCD patients. During our study window, 8 pediCD (2 NOA, 3 FR, 3 PR) received repeat biopsies which we also analyzed through scRNA-seq (see below).

Flow cytometry of the terminal ileum reveals minimal changes in leukocyte subsets in FGID vs. pediCD, and no significant differences across the pediCD spectrum

We collected terminal ileum biopsies from 14 pediCD patients and 13 uninflamed FGID patients, and prepared single-cell suspensions for flow cytometry and scRNA-seq. Biopsies from pediCD were from actively-inflamed areas adjacent to ulcerations. Biopsies from FGID were from non-inflamed terminal ileum. The epithelium was first separated from the lamina propria before enzymatic dissociation, and flow cytometric analysis was performed on the viable single-cell fraction, which recovered predominantly hematopoietic cells with some remnant epithelial cells (<20% of all cells), likely representing those in deeper crypt regions (Figure 2; Supplemental Figure 2). We utilized two flow cytometry panels, allowing us to resolve the principal lymphoid (CD4 or CD8 T cells, NK cells, B cells, innate lymphoid cells, γδ T cells, CD8αα+ IELs, pDCs) and myeloid (monocytes, granulocytes, HLA-DR+ mononuclear phagocyte) cell subsets (Supplemental Figure 2, Supplemental Table 1). From these panels, which generated 32 gates identifying cell lineages, types and subsets, only HLA-DR+ macrophages/DCs and pDCs were significantly increased in pediCD relative to FGID (Figure 2d). We also analyzed within pediCD, comparing the baseline samples of 4 NOA, 5 FR and 5 PR patients, and noted no significant differences between NOA and patients on anti-TNF, or between FRs and PRs to anti-TNF (Figure 2e). Together, this suggests that despite the substantial endoscopic, histologic, and clinical parameters that distinguish FGID and pediCD (Figure 1), the broad single-cell type composition of the terminal ileum appears minimally altered in pediCD, save for an increased frequency of pDCs and HLA-DR+ mononuclear phagocytes.

Flow Cytometry of Ileal Biopsies Does Not Reveal Significant Changes in Cell Composition in FGID vs. pediCD or across the pediCD Treatment Response Spectrum

a. Representative flow cytometry end gates for selected cell subsets (left: epithelial and hematopoietic; middle: naïve and effector T cells; right: pDCs and antigen-presenting cells) from single-cell dissociated samples from one terminal ileum biopsy for pediCD patients (see Supplemental Figure 2 for full gating strategy; numbers represent percentage of cells in each gate).

b. Fractional composition of selected cell subsets of CD45+ cells from 13 FGID and 14 pediCD patients (error bars are s.e.m).

c. Fractional composition of selected cell subsets of CD45+ cells from 4 NOA, 5 FR and 5 PR patients.

d. Fractional composition of dendritic, pDC, central memory (CM) and effector memory (EM) CD4+ and CD8+ cells from 13 FGID vs 14 pediCD patients. Dendritic cells and pDC plotted as percentage of CD45+ cells. CM/EM CD4+ and CD8+ cells plotted as percentage of total CD4+ and CD8+ cells, respectively. p < 0.05 by Mann-Whitney for pediCD versus FGID and 1-way ANOVA for pediCD cohorts).

e. Fractional composition of dendritic cells, pDCs, central memory (CM) and effector memory (EM) CD4+ and CD8+ cells from 4 NOA, 5FR, and 5 PR patients. Graphs plotted as in d.

Traditional clustering of scRNA-seq data from FGID and pediCD patients

In addition to flow cytometry, we performed droplet-based scRNA-seq on cell suspensions from the 14 pediCD and 13 FGID patients using the 10X Genomics 3’ V2 platform (Figure 1). The analyzed cell suspensions were derived from lamina propria preparations, which our flow cytometry data suggested would be composed primarily of CD45+ leukocytes, alongside a fraction of epithelial cells and stromal/vascular cells. Deconstructing these tissues into their component cells provided us with the ability to identify some of the corresponding cell types (e.g. T or B cell) and subsets (CD8αα+ IEL or CD4+ T cell) to those we identified by flow cytometry. Importantly, it also enabled us to: 1. characterize these major cell types and subsets without needing to pre-select markers, and 2. gain substantially enhanced resolution into the cell states (i.e. gene expression programs determined by co-expression of genes) within these cell types and subsets.

Following library preparation and sequencing, we derived a unified cells-by-genes expression matrix from the 27 samples, containing digital gene expression values for all cells passing quality thresholds (n=254,911 cells; Supplemental Figure 3; Supplemental Tables 2 and 3; Methods). We then performed dimensionality reduction and graph-based clustering, noting that despite no computational integration methods being used, FGID and pediCD were highly similar to each other when clustered together and visualized on a uniform manifold approximation and projection (UMAP) plot (Supplemental Figure 4a-c). We recovered the following cell types from both patient groups: epithelial cells, T cells, B cells, plasma cells, glial cells, endothelial cells, myeloid cells, mast cells, fibroblasts, and a proliferating cell cluster. We noted that the fractional composition amongst all cells of T cells, B cells, and myeloid cells was not significantly different between FGID and pediCD, similar to the flow cytometric data, and this was also the case for endothelial, fibroblasts, glial, mast and plasma cells, which were not measured through flow cytometry (Supplemental Figure 4d). This provided validation and extension of our flow cytometry data documenting that the broad cell type composition of FGID and pediCD is not significantly altered, despite highly-distinct endoscopic and histologic diseases. Based on this joint clustering and annotation of top-level cell types, we then performed differential expression testing identifying significant up- and down-regulated genes across cell types (Supplemental Figure 4e; Supplemental Table 4). Within myeloid cells we identified some of the most significantly upregulated genes in pediCD versus FGID to be CXCL9 and CXCL10, canonical IFNγ-stimulated genes, and S100A8 and S100A9 which form the biomarker fecal calprotectin (Supplemental Figure 4f)(Leach et al., 2007; Ziegler et al., 2021). Within epithelial cells we identified that APOA1 and APOA4 were amongst the most significantly downregulated genes, and correspondingly REG1B, SPINK4 and REG4A were amongst the most significantly upregulated indicating tradeoffs between lipid metabolism and host defense in pediCD versus FGID (Supplemental Figure 4f) (Haberman et al., 2014). In T cells, we noted that the cytotoxic genes GNLY and GZMA were amongst the most significantly upregulated in pediCD, with almost no genes downregulated (Supplemental Figure 4f).

We then systematically re-clustered each broad cell type, identifying increasing cellular heterogeneity. Given that we detected changes in the frequency of HLA-DR+ macrophages/dendritic cells and pDCs between pediCD and FGID by flow cytometry, we initially focused on the myeloid cell type sub-clustering, containing dendritic cells, macrophages, monocytes, and pDCs (Supplemental Figure 4g). Working within this analysis paradigm revealed that a traditional clustering approach had difficulty identifying the boundaries of clusters, and whether a cluster composed primarily of pediCD rather than FGID cells represented a unique cell subset, or a cell state overlaid onto a core cell subset gene expression program (Supplemental Figure 4g, Methods). These initial findings also raised the possibility that subjective analytical decisions in this manual joint clustering approach might obscure some of the unique biology of FGID and pediCD. For instance, this analytical approach could lead to hybrid clusters, informed by cells from both FGID and pediCD, that according to recent Human Cell Atlas-scale efforts in the lung, may limit discovery of rare but critical disease-specific cell types, subsets, or states (Sikkema et al., 2022).

ARBOL automated iterative tiered clustering

In order to approach the analytic challenge from a more principled direction, we made four key changes to our workflow. First, as FGID represents a non-inflamed condition and pediCD and inflamed state, we proceeded to analyze FGID and pediCD samples separately to define corresponding cell type, subset, and state clusters and markers in order to respect the underlying disease biology and maximize the potential for high-quality annotation of cell subsets. Second, to minimize analyst-to-analyst variability in the choice of clustering parameters, which can significantly under- or over-cluster data, we developed ARBOL (https://github.com/jo-m-lab/ARBOL). ARBOL is an automated clustering approach—built with conventional Seurat functions—to maximize the silhouette score at each tier of iterative sub-clustering and stop sub-clustering when a specific granularity is reached (Supplemental Figure 5a,b; Methods), Third, we systematically generated descriptive names for cell types and subsets together with differentiating marker genes. Fourth, we accounted for the number and diversity of patients which compose each cluster using Simpson’s Index of Diversity. This enabled us to focus our study on clusters that, despite being rare, were reproducibly found in the majority of patients (Simpson, 1949). Using ARBOL, each tier of analysis is typically under-clustered relative to traditional empirical analyses, but the automation proceeds through several more tiers (typically 4 to 5) until stop conditions (e.g. cell numbers and differentially expressed genes; Methods) are met. Intriguingly, this number of tiers was also manually selected in the HCA expert-annotated lung cell atlas, where reliable and robust cell clusters were found in the 0.05% cell frequency range (Sikkema et al., 2022).

We then inspected all outputs (182 FGID and 425 pediCD clusters) and provided descriptive cell cluster names independently for FGID and pediCD (Supplemental Figures 5b and 6). We also focused at this stage on flagging putative doublet clusters or clusters where the majority of differentially expressed genes which triggered further clustering consisted of known technical confounders in scRNA-seq data (e.g. mitochondrial, ribosomal, and spillover genes from cells with high secretory capacity) yielding a final number of 118 FGID and 305 pediCD clusters (Supplemental Figure 5b) (Smillie et al., 2019). Importantly, this quality-control step could often be missed by traditional scRNA-seq analysis, since these analysis paradigms retain such cells within higher-tier clusters, yet their inclusion of these cells would confound both differential composition and expression analyses if not identified. We note this clustering method represents a data-driven approach, though it may not always reflect a cellular program or transcriptional module of known biological significance or stability. Nevertheless, we highlight that many of these small clusters (∼100 cells; 0.1% of total cells) which may represent transient cell states, are found across all FGID or pediCD patients surveyed, thus suggesting a level of consistency across diseases, and across genetically and environmentally unrelated individuals.

Two comprehensive atlases of FGID and pediCD

To generate the final dendrograms for the FGID and pediCD cell atlases, we hierarchically clustered all end cell states, and performed 1 vs. rest within-Tier 1 clusters (i.e. broad cell types) differential expression to provide systematic names for cells. The naming heuristic was based on the cell type classification, cell subset classification with literature-based markers, and two genes using a rank-score function (Figures 3 and 4; Supplemental Figure 6; Methods). We provide complete gene lists at three levels of comparison: cell types (1 vs. rest across all cells), subsets (1 v. rest across all cell types), and states (1 vs. rest within-Tier 1 cell-type) in Supplemental Tables 5-10. More specifically, to select marker genes for naming in a data driven manner, we used 1 vs. rest within-cell-type differential expression (Supplemental Tables 7 and 10; Wilcoxon, Bonferroni adjusted p<0.05), and a rank-score function (-log(sig+1) * avg_logFC * (pct.in / pct.out)) to prioritize the top 2 marker genes. For example, using this ranking system, we identified CCL3 and CD160 as two genes significantly enriched in one NK cluster (adj. p-value = 0, expression within-cluster >40% cells positive and in other Tier 1 T cells <6%). This resulted in a final name for this cluster of CD.NK.CCL3.CD160. We repeated this process for all FGID and pediCD cells within Tier 1 B cell, endothelial, epithelial, fibroblast, plasma cell, myeloid cell, mast cell, and T cell identified clusters, and provide systematically generated names for all (Supplemental Figure 5; Supplemental Tables 11 and 12).

A Comprehensive Cell Atlas of Terminal Ileum in Non-Inflammatory FGID

a. tSNE of 99,488 single-cells isolated from terminal ileal biopsies of 13 FGID patients. Colors represent major cell type groups determined via Louvain clustering with resolution set by optimized silhouette score.

b. tSNE as in a with individual patients plotted. For specific proportions please see Supplemental Figure 4.

c. tSNE of each major cell type which was used as input into iterative tiered clustering (ITC).

d. End clusters determined by ARBOL of complete FGID data set were hierarchically clustered on the median expression of 4,445 pairwise differentially expressed genes, using complete linkage and distance calculated with Pearson correlation, between each end cell cluster. Simpson’s Index of Diversity represented as 1-Simpson’s where 1 (black) indicates equivalent richness of all patients in that cluster, and 0 (white) indicates a completely patient-specific subset. Numbers represent the number of cells in that cluster. Names of subsets are determined by Disease.CellType.GeneA.GeneB as in Methods.

e. Dot plot of 2 defining genes for each cell type. Dot size represents fraction of cells expressing the gene, and color intensity represents binned count-based expression level (log(scaled UMI+1)) amongst expressing cells. All cluster defining genes are provided in Supplemental Tables 5 to 7.

A Comprehensive Cell Atlas of Terminal Ileum in pediCD

a. tSNE of 124,054 single-cells isolated from terminal ileal biopsies of 14 pediCD patients. Colors represent major cell type groups determined via Louvain clustering with resolution set by optimized silhouette score.

b. tSNE as in a with individual patients plotted. For specific proportions please see Supplemental Figure 4.

c. tSNE of each major cell type which was used as input into iterative tiered clustering (ITC).

d. End clusters determined by ARBOL of complete pediCD data set were hierarchically clustered on the median expression of 1,844 pairwise differentially expressed genes, using complete linkage and distance calculated with Pearson correlation, between each end cell cluster. Simpson’s Index of Diversity represented as 1-Simpson’s where 1 (black) indicates equivalent richness of all patients in that cluster, and 0 (white) indicates a completely patient-specific subset. Numbers represent the number of cells in that cluster. Names of subsets are determined by Disease.CellType.GeneA.GeneB as in Methods.

e. Dot plot of 2 defining genes for each cell type. Dot size represents fraction of cells expressing the gene, and color intensity represents binned count-based expression level (log(scaled UMI+1)) amongst expressing cells. All cluster defining genes are provided in Supplemental Tables 8 to 10.

From the 99,488 cells profiled from 13 FGID patients, we recovered 12 Tier 1 clusters, representing the main cell types found in the lamina propria and remnant epithelium contained in an ileal biopsy, which we display on a t-stochastic neighbor embedding (t-SNE) plot colored by cluster identity (Figure 3a; Supplemental Figure 5b). From the 124,054 cells profiled from 14 pediCD patients, we recovered 12 Tier 1 clusters which we also display on a t-SNE plot colored by cluster identity (Figure 4a; Supplemental Figure 5b). Distinct from FGID, Paneth cells clustered separately at Tier 1 in pediCD, while glial cells were now found within the fibroblast Tier 1 cluster. Inspecting each individual patient’s contribution to each broad cell type, we noted that all patients contributed to all Tier 1 clusters in both FGID and pediCD (FGID: Figure 3b; pediCD: Figure 4b; Supplemental Figure 4c,d; Methods Supplemental Table 13). As patient identity did not factor into ARBOL stop conditions, we then calculated Simpson’s Index of Diversity for each of the end clusters (Figure 3d; Supplemental Figure 5). Although low diversity clusters may still reflect important biology for individual patients, we comment more extensively on clusters with high patient diversity. In pediCD, most end clusters are conserved across multiple patients, while only 16/305 clusters were single-patient clusters (Figure 4d; Supplemental Figure 5). We note that in pediCD, relative to FGID, a higher fraction of cell clusters exhibited lower patient diversity. In the Methods section, we provide full explanations for how names for subsets and clusters were defined for B cells, myeloid cells, T/NK/ILC’s, epithelial cells, endothelial cells, fibroblasts, mast cells and plasma cells, and include links to corresponding Supplemental Tables containing full supporting gene lists and data visualization. We share portal links in Data Availability.

Using this analytical workflow, we present two comprehensive cellular atlases, separately mapping FGID (Figure 3) and pediCD (Figure 4). The pediCD atlas enabled us to nominate cell states associated with disease severity and treatment outcomes in this disease (Figures 5 and 6), identify histological relationships between epithelial/myeloid/lymphoid cells (Figure 7), and understand how anti-TNF pushes the pediatric treatment-naïve ecosystem towards an adult tissue state (Figure 8).

A Collective Cell Vector in pediCD Reveals Predictive Axes of Disease Trajectory and Treatment Response

a. Spearman rank correlation heatmap of principal components calculated from the frequencies of each end cluster per main cell type together with clinical metadata. Correlation is represented by both the intensity and size of the box and those which are FDR < 0.05 have a bounding box.

b. A PCA plot of each patient’s end cell cluster composition as determined by ARBOL for T/NK/ILC, from the pediCD dataset where the end cell clusters frequency is calculated amongst all cell types and colors represent anti-TNF response. (inset highlights the specific correlation between PC2 of the T, Myeloid, Epithelial cell frequency analysis with anti-TNF response). We term this principal component “pediCD-TIME”.

c. Cell cluster frequencies of the parent cell type found to be significant by Mann-Whitney U test between selected clusters (see Supplemental Figure 7 for all graphs; Supplemental Table 15).

d. Heatmap showing cell frequencies per patient of most positive and negative cell subsets of PC2 from PCA performed on T/NK/ILC, myeloid and epithelial cell subsets (Supplemental Table 16). Cell subsets are sorted by PC2 score, and patients were sorted by anti-TNF response. Heatmap is not normalized and displaying the log counts-per million of each cell subset normalized per cell type. *Patient p022’s response category changed from FR to PR after database lock in December of 2020. No other patient’s categorization has changed.

ARBOL analysis of joint FGID and pediCD T/NK/ILCs, Epithelial and Myeloid cells identifies high-diversity clusters that drive disease severity signature

a. A principal component analysis (PCA) plot of each patient’s end cell cluster composition as determined by ARBOL for T/NK/ILC, Myeloid, and Epithelial cells from FGID and pediCD patients where the end cell clusters frequency is calculated amongst total cells and colors represent disease state and treatment response.

b. The Spearman rank correlation between PC1 of the joint T/NK/ILC, Myeloid, and Epithelial cell frequency analysis with anti-TNF response.

c. A UMAP plot of 7,366 activated T/NK/ILC cells from FGID (colored by disease) and pediCD (colored by treatment response) data sets.

d. Hierarchical clustering of activated T/NK/ILC cells from FGID and pediCD data set with input clusters determined based on results of ARBOL, and performed on the median expression of all genes, using complete linkage and distance calculated with Pearson correlation, between each end cell cluster. Simpson’s Index of Diversity represented as 1-Simpson’s where 1 (black) indicates equivalent richness of all patients in that cluster, and 0 (white) indicates a completely patient-specific subset. Numbers represent the number of cells in that cluster. Names of subsets are determined by Disease.CellType.GeneA.GeneB as in Methods, and names in red indicate strongest PR-PC1-negative loadings. Pie charts represent the fractional composition of FGID or pediCD cells (note the initial proportions for reference).

e. Top JOINT-PC1-negative clusters are both (left) high diversity and (right) almost exclusively found in pediCD patients.

Histological and scRNA-seq analyses indicate CD4+ T cell infiltration into the epithelium of pediCD samples

a. Representative (left) multiplex immunofluorescence of B cells (CD20), T cells (CD3), cytotoxic T cells (CD8), regulatory T cells (FOXP3), myeloid cells (CD68) and epithelium (panCK) which was then (right) analyzed using an automated classifier mask to divide sections into epithelium, lamina propria and glass areas for cellular quantification.

b. Representative CD and FGID sections stained with indicated markers, inset focuses on T cells within epithelium. Quantification of 8 pediCD and 10 FGID participants is represented as a row z-scored hierarchically-clustered heatmap for total number of indicated cells per mm2 of whole section, epithelium, or lamina propria.

c. Cells per mm2 of epithelium for CD3+ cells, CD3+CD8- cells (CD4 T cells), and CD3+CD8- FOXP3- (effector CD4 T cells) for pediCD and FGID participants.

d. Spearman rank correlation heatmap of the counts-per-million for each of the top 25 clusters defining pediCD-TIME positive (NOA-associated) and pediCD-TIME negative (PR-associated) vectors. Correlation is represented by both the intensity and size of the box and those which are FDR < 0.05 have a bounding box

Anti-TNF treatment shifts the pediCD cellular ecosystem towards adult treatment-refractory disease

a. GSEA analysis showing the ranks of 92 PREDICT markers (markers of top 25 cell states associated with disease severity and treatment outcomes) in bulk RNA sequencing of illeal or colonic mucosa of two other treatment-naïve cohorts (pediatric RISK cohort, n = 69, adult E−MTAB−7604 cohort, n = 43) comparing pediCD patients who did or did not respond to anti-TNF therapy. P-value is estimated based on an adaptive multi-level split Monte-Carlo scheme.

b. UMAP plots of 275,544 cells from PREDICT ileal biopsies (Baseline: 107,432 cells; Repeat: 47,796 cells) or GIMATS anti-TNF treated ileal resections (GIMATS Uninflamed: 61,965; GIMATS Inflamed 58,351) integrated using Harmony (batch = 10x version).

c. A UMAP plot from cells in (b) colored by major cell type.

d. A PCA plot (left) or UMAP plot (right) of each patient’s end cell cluster composition as determined by ARBOL with Harmony integration (batch = 10x version) for T, B, Myeloid, Fibroblast, Plasma from the treatment-naïve PREDICT pediCD samples (n=13), repeat PREDICT biopsies (n=2 NOA, 6 on-anti-TNF), and adult on-treatment GIMATS biopsies (n=22) where the end cell clusters frequency is calculated amongst total cells and colors represent severity and cohorts are shapes.

e. A UMAP trajectory of the CPM table underlying (d).

f. The top 50 varying (yellow) cell clusters along the pseudotime trajectory as determined by Moran’s I test (q-value < 0.000029).

g. Cells per million of significantly varying T/NK/ILC clusters where individual points represent patient biopsies colored by severity and cohorts as shapes (as in d)

Clinical variables and cellular variance that associates with pediCD severity

As the pediCD atlas was curated from treatment-naïve diagnostic samples, we were able to interrogate the data to test if overall shifts in cellular composition, specific cell states, and/or gene expression signatures underlie clinically-appreciated disease severity and treatment decisions (NOA vs. FR/PR), and those that are further associated with response to anti-TNF therapies (FR vs. PR). Here, we leveraged the detailed clinical trajectories collected from all patients as the ultimate functional test: resolving how cellular composition and cell states anticipate disease and treatment outcomes.

In order to capture the overall principal axes of variation explaining changes in cellular composition, we calculated the fractional composition of all 305 end cell clusters in pediCD within its parent cell type (“per cell type”), or within all cells (“per total cells”), and performed a principal component analysis (PCA) over both of these sample x cell cluster frequency tables (Supplemental Table 14) (Mathew et al., 2020). We then used the PC1 (13.4% variation “per cell type” and 13.5% variation “per total cells”) and PC2 (12.7% variation “per cell type” and 11.8% variation “per total cells”) values for each sample as numerical variables which we correlated with clinical metadata (Figure 5a, r by Spearman-rank). Amongst the clinical variables, we noted strong correlation between Initial wPCDAI and CRP (r=0.83, FDR<0.05), and moderate correlation between Initial wPCDAI and either anti-TNF within 90 days (r=0.65) or anti-TNF response coded as NOA (0) FR (1) and PR (2) (r=0.49). The pediCD atlas also enabled us to correlate the transcriptome with initial disease severity and treatment response: Thus, for PC1- “per total cells”, we identified strong correlations with anti-TNF treatment within 90 days (r=-0.76), and moderate correlation anti-TNF_NOA_FR_PR (r=-0.58; Methods).

Discovery of pediCD-TIME: a collective cell vector that anticipates CD clinical severity and response spectrum

In order to understand if multiple cell types, acting in a concerted manner, were predominantly driving the association with clinical disease severity at initial presentation of pediCD, we then further deconvoluted the overall PCA on 305 end clusters and performed PCA over each cell type’s fractional composition of end clusters individually (B cells: 33 clusters, Endothelial: 18 clusters, Epithelial: 68 clusters, Fibroblast 45 clusters, Myeloid: 54 clusters, T/NK/ILCs: 57 clusters), and correlated the first two PCs (all PC1s and PC2s each accounted for >13% variance) with all of the clinical variables (Supplemental Table 14). The PCs derived from T/NK/ILC cells, myeloid cells, and epithelial cells were all moderately correlated with anti-TNF_NOA_FR_PR status (r>0.49) individually, and had higher values than the other cell types; therefore, we asked if a PCA-based metric considering all three cell types would synergistically capture both disease severity and treatment response. When we calculated the PCA accounting for frequencies of each cell subset of T/NK/ILC cells, myeloid cells, and epithelial cells amongst all cell types together, we found strong correlation for PC2 with both anti-TNF within 90 days (r=-0.83) and anti-TNF-NOA_FR_PR status (r=-0.87, FDR<0.05) (Figure 5a,b). This represented the two strongest correlations of any variable we tested with anti-TNF treatment and response status. This “PC2-T/NK/ILC/Myeloid/Epithelial” is further referred to as ‘pediCD T cell, innate lymphocyte, myeloid and epithelial’ (pediCD-TIME).

To further deconstruct pediCD-TIME, we used frequency-based statistics and analysis of PC loadings. We first identified which cell clusters accounted for the most significant changes in relative frequency based on the relative frequency of an end cell cluster within its parent cell type, noting limitations related to dissociation-induced biases (Discussion) (Gomariz et al., 2018). We performed a Fisher’s exact test between NOA vs. FR; NOA vs. PR; or FR vs. PR, and then performed a Mann-Whitney U test to highlight specific clusters (Fig. 5c; Supplemental Figure 7; Methods). As a comparison to differential expression within a cell type or subset, we highlight the power of ARBOL to deconstruct cell clusters that are often unified through strong cell-state signatures, such as proliferation, into identifiable, diverse, and disease-associated cell subsets (Supplemental Figure 8; Methods). This cluster-level analysis identified concerted changes when comparing FRs and PRs to NOAs, especially within the T/NK/ILC compartment, with PRs relative to NOAs having additional significant changes within the myeloid compartment (Fig. 5c; Supplemental Figure 7; Methods).

As noted above, higher disease severity was inversely related with the PC loadings in pediCD-TIME. Amongst the top negative PC loadings for pediCD-TIME (Fig. 5a,b), enriched in FRs—and further enriched in PRs—compared to NOAs, included both helper and cytotoxic T cell clusters (CD.T.MAF.CTLA4; CD.T.CCL20.RORA; CD.T.GNLY.CSF2), NK cell clusters (CD.NK.GNLY.FCER1G; CD.NK.GNLY.IFNG; CD.NK.GNLY.GZMB); proliferating T cells and NK cells (CD.T.MKI67.FOXP3; CD.T.MKI67.IFNG; CD.T.MKI67.IL22; CD.NK.MKI67.GZMA), and monocytes, macrophages, DCs and pDCs (CD.cDC2.CD1C.AREG; CD.Mac.C1QB.CD14; CD.Mono.CXCL3.FCN1; CD.pDC.IRF7.IL3RA; CD.Mono/Mac.CXCL10.FCN1) (Figure 5d; Supplemental Table 16). The top positive loadings for pediCD-TIME encompassing the NOA-enriched clusters included several epithelial cell subsets such as Tuft cells (CD.EC.GNAT3.TRPM5) and those with specialized metabolic features including retinol-binding, bile binding and export, fatty-acid and cholesterol metabolism, fructose and glucose metabolism, starch metabolism, glutathione metabolism, sulfation, and the terminal degradation of peptides (CD.EC.RBP2.CYP3A4; CD.EC.FABP6.PLCG2; CD.EC.FABP1.ADIRF; CD.EC.GSTA2.TMPRSS15) (Figure 5d; Supplemental Table 16) (Lampen et al., 2000; Mårtensson et al., 1990; Martínez-Augustin and de Medina, 2008; Sullivan et al., 2021; Wen and Rawls, 2020). Furthermore, clusters also enriched in NOA-pediCD-TIME such as CD.EC.ADH1C.RPS4Y1 and CD.EC.ADH1C.GSTA1, clustered in a separate branch together and expressed several enzymes responsible for steroid hormone and dopamine biosynthesis (Figure 4d, 5d) (Cima et al., 2004; Magro et al., 2002). Importantly for the regenerative potential of the epithelium, CD.EpithStem.LINC00176.RPS4Y1 were also defining of the NOA-pediCD-TIME direction.

In order to determine the relative contributions and stability of each cell cluster to the pediCD-TIME vector, we performed leave-one-out cross-validation (LOOCV) on the top-weighted pediCD-TIME clusters based on their frequency within total cells (Supplemental Figure 9a). This identified the stability of cell clusters such as CD.T.MAF.CTLA4, CD.Mono.CXCL3.FCN1, CD.T.GNLY.CSF2 and other top PR-pediCD-TIME-negative subsets. Furthermore, we also re-calculated PC weights as frequencies of clusters within their major cell type, with the standard deviation of PC loadings stabilizing even further (Supplemental Figure 9b). This suggests that compositional scRNA-seq analysis provides more robust signatures within major cell types than amongst all total cells, and both methods can recover a disease trajectory.

Together, these discoveries underscore the multiple collective changes in the composition and/or state of T/NK/ILC cells, myeloid cells, and epithelial cells at pediCD diagnosis that could stratify pediCD patients by disease severity, and may influence anti-TNF responsiveness.

Cluster-based and cluster-free approaches contextualize pediCD with FGID

As we had generated independent cellular atlases for FGID and pediCD to mitigate aspects of disease batch effects obscuring the ability to resolve cell clusters, we next sought to compare pediCD and FGID cell subsets using label-retaining (Random Forest, ARBOL) and label-free (diffusion maps) methods. We applied these models to each cell type individually, and here focus our discussion on Myeloid cells and T/NK/ILC cells as two cell types prominently associated with pediCD disease severity (Supplemental Figures 10-12). As newer methods are developed, more refined integration is likely to be possible. Comparing across myeloid cells between pediCD and FGID, we could identify strong correspondence of specific cell subsets such as cDC1s or pDCs (Supplemental Figure 10a). We also identified strong correspondence between several cDC2 clusters. We identified a gradient of monocyte and macrophage correspondence of 31 clusters in pediCD to 2 FGID clusters, likely reflective of inflammatory monocyte to macrophage differentiation in pediCD (Blériot et al., 2020; Dutertre et al., 2019; Guilliams et al., 2018). For T/NK/ILC cells, we identified more discrete patterns relative to Myeloid cells based on comparison of the RF result. Within the two FGID cytotoxic T cell clusters, we identified correspondence by 18 pediCD clusters, representing ILC3s, and cytotoxic NK cells and T cells (Supplemental Figure 11). The cluster of naïve T cells in FGID had correspondence with the majority of pediCD non-cytotoxic T cell clusters, illustrating a substantial activation and specialization to several discrete T cell states that were specific for pediCD.

We next performed an analysis over a shared gene expression space of FGID and pediCD of the monocytes/macrophages (Supplemental Figure 13) and T/NK/ILCs (Supplemental Figure 14). Within FGID monocytes/macrophages, we identified that the majority of clusters occupied the periphery of the UMAP space, including chemokine-expressing clusters (FG.Mac.CCL3.HES1; FG.Mac.CXCL8.IL1B) and metabolic clusters (FG.Mac.APOE.PTGDS) (Supplemental Figure 13a,c). This was in stark contrast to the pediCD monocytes/macrophages, where we identified that now many of the clusters occupied the central region of the UMAP (Supplemental Figure 13a,c). We highlight several of the clusters that are significantly different in frequency between the pediCD groups which were found in this central region (Supplemental Figure 13b), and how NOA, FR and PR pediCD clusters had significantly different distributions within this space as measured by the Hellinger Distance (Supplemental Figure 13c-e; Methods). Of note, the frequency of TNF+ macrophages and its expression level of TNF was significantly increased in FRs relative to all other groups (Supplemental Figure 13f; permutation test shuffling anti-TNF response variable, FR had significantly more TNF+ cells than expected by chance with p approximating 0). Despite the fine-grained tiered clustering approach used, the majority of clusters had high Simpson’s Diversity indices representing cell states found in several patients (Supplemental Figure 13g).

Within T/NK/ILCs, we identified that FGID cells were more uniformly mixed with the pediCD cells relative to monocytes/macrophages (Supplemental Figure 14a). FGID cells occupied naïve and quiescent states, showed some signatures of activation, and also specialization towards helper and cytotoxic states (Supplemental Figure 14a,c) (Sallusto et al., 1999). The most notable changes in the Hellinger Distance distribution occurred between FGID and FR, rather than between FGID and PR (Supplemental Figure 14d,e). Similar to the monocytes/macrophages, the main area which gained density with increased disease severity was the central region: characterized by proliferation of several clusters increased in frequency within FRs and PRs, including T cells such as CD.T.MKI67.FOXP3 and CD.T.MKI67.IL22 (Supplemental Figure 14b-d). Taken together with several cell clusters associated with pediCD proliferation overlapping with existing areas found in FGID, this indicates activation and more extreme diversification from existing T cell states driving the T cell clustering that defined pediCD. This is distinct from the recruitment and failed differentiation towards homeostatic cell states between FGID and pediCD that we discovered in monocytes/macrophage clusters. Both joint projections confirm and extend our random forest correspondence probabilities, and provide a clustering-independent view of how cellular density is shifted between FGID and pediCD (Dann et al., 2022).

Joint ARBOL atlas of activated T/NK/ILC’s in FGID and pediCD

Building on our Random Forest and Hellinger Distance distribution analyses, we sought to formally cluster pediCD and FGID T/NK/ILC, Myeloid, and Epithelial cells together to place the pediCD patients within context of FGID, taking advantage of the high-resolution view afforded by ARBOL (Fig. 6a). We tested whether performing ARBOL on a joint FGID and pediCD dataset would also discern a disease severity vector. We identified that JOINT-PC1 in this analysis captures a disease severity vector, with the positive direction enriched in FGID patients, and the negative direction enriched in FR and PR pediCD patients (Fig. 6b; r= -0.74, p<0.0001). We focused on a branch of 7,366 activated T/NK/ILCs, annotated all end clusters as T, NK or ILC (Supplemental Figure 6), and selected 2 genes from the top-10 rank-scored list (Methods) that provide a descriptive name for each cluster (Fig. 6c,d). We visualized at each split of the binary tree representation the fractional composition of cells weighted towards FGID or pediCD cells.

Our tree recovered primary splits of memory T cells, cytotoxic cells, helper cells, and intraepithelial cells, with proliferative cells interspersed within these groups. Intriguingly, this analysis reinforces our flow cytometric and initial two-tiered manual clustering approach whereby up to a certain level of the tree, FGID and pediCD do not show significant differences in composition. However, when following the additional splits leading to each end cluster, we now note several clusters which are pediCD or FGID enriched. Furthermore, despite this disease-enrichment, the end clusters are still high in patient diversity, with the majority falling above 0.5 Simpson’s Index (Figure 6d,e). Importantly, our top-ranked PR-PC1-negative T cells contributing to the disease severity vector are all high diversity clusters even when clustered jointly with FGID (Fig. 6; Supplemental Figure 15). We again recover a majority of proliferating T and NK cell clusters contributing to the JOINT-PC1 disease severity vector (NK.MKI67.CD38, T.MKI67.FABP5, T.MKI67.IL17A, T.MKI67.IL26, T.MKI67.FOXP3, and others).

While further experiments would need to be conducted to isolate and determine the stable or transient nature of each cell cluster, their conservation across multiple patients from distinct genetic and environmental backgrounds suggests that these end clusters may be reliably found across most patients sampled within a disease category, and that the primary disease-associated signatures remain robust when placed in context of another pediatric gastrointestinal disease. As we had now generated and annotated three atlases: FGID, pediCD, and joint FGID/pediCD, we took the opportunity to calculate how considering atlases separately or jointly influenced the coherence of each end cluster (Supplemental Figure 15; Methods). We anticipate that as atlases scale in cell number, patient number, cohort number, and meta-analyses, that multiple methods of clustering will be important to fully-resolve the cellular heterogeneity present within each sample, while retaining community-level labels for facile integration and comparison for meta-analyses.

Pair-Wise Correlation of Cell Clusters and Histology

In order to provide spatial quantification for key cell types in our disease signature, we also utilized multiplex immunofluorescence to identify the absolute numbers and locations of B cells, cytotoxic, helper and regulatory T cells, and macrophages in FGID and pediCD samples. Using an automated mask for epithelial and lamina propria identification, we were able to resolve the location of these cell types within the whole section or within the epithelial or lamina propria layers of the tissue (Figure 7a). We identified no significant differences in the whole sections for total numbers of T cells (CD3+), B cells (CD20+), myeloid cells (CD68+), and cytotoxic T cells or regulatory T cells between pediCD (n=8) and FGID (n=10) participants (Figure 7b). Strikingly, we did find a significant difference in total CD3+ cells within the epithelium, and specifically of CD3+CD8-FOXP3- (CD4+ effector T cells) using our multiplex immunofluorescence panel (Figure 7c). This indicates that CD4+ cells infiltrating into the epithelial layer may have an early and outsized role in driving pediCD.

In order to understand the impact of which CD4+ T cell states that were associated with anti-TNF response may be influencing specific epithelial states, we correlated proliferating T and NK cell clusters with epithelial cell clusters (Figure 7d). We found that CD.T.MKI67.FOXP3 were strongly positively associated with CD.Goblet.RETNLB.ITLN1 and CD.EC.NUPR1.LCN2 secretory cell states. Conversely, we found that CD.T.MKI67.IL22 were significantly negatively correlated with CD.EC.GNAT3.TRPM5 and EC.FABP1.ADIRF cells, and that CD.T.CCL20.RORA were significantly negatively correlated with EC.ADH1C.GSTA1 and EpithStem.LINC00176.RPS4Y1. This indicates the potential for CD4 T cells present within the epithelial layer to significantly alter epithelial cell states away from nutrient sensing and homeostatic metabolic function in pediCD.

Contextualization to published Crohn’s Disease Studies

To determine whether the pediCD-TIME disease severity gene signature that we discovered in the PREDICT study can be found in other cohorts with published bulk RNA-seq data, we selected the top 92 markers of the 25 cell states associated with disease severity and treatment outcomes (Methods; Supplemental Table 17) and performed a gene-set enrichment analysis (GSEA) (Figure 8a; Supplemental Figure 16). We used the cell clusters ranked by cell frequency of total cells, as this would be most representative of bulk-analyzed tissue. In the PREDICT cohort, we were able to validate the 92-gene signature as significantly enriched within FR and PR patients relative to FGID or NOA (all p<9.3×10-05). These 92 genes were also enriched in PR relative to FR patients (p<4.3×10-05). Notably, in the two independent treatment-naïve cohorts that we analyzed (the pediatric RISK cohort, n = 69, and the adult E−MTAB−7604 cohort, n = 43) this 92-gene signature was significantly enriched in illeal, but not colonic, mucosal biopsies from patients who did not respond to anti-TNF therapy compared to those who responded (Figure 8a; p= 0.0046 vs. p=0.78, respectively) (Kugathasan et al., 2017; Verstockt et al., 2019). Thus, these 92 genes (including TNFAIP6, GZMB, S100A8, CSF2, CLEC4E, S100A9, IL1RN, FCGR1A, CLIC3, CD14, PLA2G7, FAM26F, IL3RA, NKG7, IL32, CCL3, OLR1, LILRA4, APOC1, MYBL2 and others; Supplemental Table 17) nominate a signature of anticipatory markers of anti-TNF therapy outcome in newly-diagnosed patients and are validated in one internal (bulk RNA-seq from PREDICT patients) and two external (RISK and E-MTAB-7604) bulk RNA-seq cohorts.

Anti-TNF treated repeat pediCD biopsies are closer to adult treatment-refractory CD

We sought to test how anti-TNF treatment shifts the cellular ecosystems of the pediatric ileum. The availability of repeat biopsy specimens from PREDICT patients enabled us to perform a joint ARBOL containing treatment-naïve pediCD biopsies (n=14), 8 follow-up endoscopy samples (from 2 NOA and 6 on-anti-TNF PREDICT patients), and 11 adult CD patients from the (Martin et al., 2019) study which identified the GIMATS (T, B, Myeloid, Fibroblast Plasma) cellular module (Figure 8b).

Due to the utilization of several 10X 3’ kit versions, we added the integration method Harmony into the ARBOL pipeline (Figure 8b,c; Methods). Of note, pediCD atlas curated clusters were preserved with a similar distribution of Shannon entropy in the integrated atlas compared to the joint atlas (Supplemental Figure 15).This analysis illustrated that we continue to recover a gradient of disease severity based on our patients’ cell clusters from NOA, to untreated FR and PR. Intriguingly, the repeat anti-TNF treated samples begin to occupy the area between the baseline PREDICT samples and the on-treatment GIMATS adult samples (Figure 8d; diamonds are repeat samples). Meanwhile, the two NOA patients with repeat biopsies remain at the beginning of the trajectory overlaid next to their corresponding baseline samples.

From the resultant integrated ARBOL, we calculated a pseudotime trajectory of patients informed by cell clusters. By calculating the cell clusters most auto-correlated with the pseudotime trajectory, this revealed that T/NK/ILC, Myeloid, and Epithelial cell clusters (i.e. the same cell types defining the pediCD-TIME signature) were enriched in those tracking with the trajectory (Figure 8e,f). By using the automated naming function within ARBOL, we were able to provide identities for these cell clusters, and present those for T/NK/ILC cells (Figure 8g). Our analysis revealed that T.KLF2.FAM65B.CCR7.SELL (expressing markers of naïve T cells), T.GZMK.CCL5.HLA-B.TRAC (expressing markers of CD8 cytotoxic T cells) and T.S100A4.CCR6.TMSB4X.PFN1 (expressing markers of Th17 cells) were significantly decreased with anti-TNF treatment in pediatric and adult patients. Correspondingly, T.IL7R.TNFAIP3,KLRB1 (with signatures of TNF activation), T.CCL5.CD8A.CD8B.TXNIP (effector CD8 T cells), and T.KIT.PCDH9.LINC00299.AREG; T.LST1.AREG.CXCL2.LINC00299 (innate lymphoid cell clusters) increased. This analysis identifies those clusters within the terminal ileum which are most associated with compensation to TNF inhibition, and illustrates that while our NOA patients exhibited remarkable consistency in the pseudotime location of their biopsies within this trajectory, anti-TNF treatment consistently “pushed” patient’s cells closer to the location of the anti-TNF refractory disease observed in adult CD.

Discussion

This study addresses a critical unmet need in the fields of IBD and systems immunology: the creation of an atlas of newly-diagnosed untreated diseased tissue, coupled with detailed clinical follow-up to link diagnostic cell types and states with disease trajectory. This is especially true for GI autoimmune disease, and other diseases which affect tissues not easily accessible without operative or endoscopic intervention, and where tissue-specific immune pathology dictates disease severity and trajectory. Likewise, cross-sectional studies, as have been the norm for most previous scRNA-seq studies of IBD, are not able to overlay disease trajectory and treatment response onto the topography of a complex multi-cellular atlas, thus limiting the mechanistic and predictive inferences that can be drawn from the generated atlas (Corridoni et al., 2020b, 2020a; Drokhlyansky et al., 2019; Elmentaite et al., 2020; Huang et al., 2019; Kinchen et al., 2018; Martin et al., 2019; Parikh et al., 2019; Smillie et al., 2019; Uzzan et al., 2022).

Furthermore, mouse models of CD, and of IBD more broadly, may not be the most appropriate models for understanding treatment resistance in pediCD (Neurath, 2019). To surmount these limitations, we created a prospective clinical study, and enrolled patients requiring a diagnostic biopsy for possible IBD, prior to diagnosis. This allowed us to capture a tremendously valuable control group: those patients with FGID, who experience GI symptoms, but for whom the endoscopy does not reveal evidence of GI inflammation or autoimmunity. These uninflamed controls served as a critical comparator to contextualize the evidence of immune pathology that we observed in patients with pediCD across the severity spectrum. With these detailed clinical phenotypes as our foundation, we developed an automated iterative tiered clustering algorithm for scRNA-seq data, ARBOL, which defined pediCD-TIME, a vector of T cells, myeloid cells and epithelial cells that stratifies both pediCD disease severity and response to treatment. We also relate our treatment-naïve samples to repeat samples from PREDICT patients on anti-TNF, and to an adult on-treatment Crohn’s scRNA-seq atlas, identifying a continuum between pediatric and adult Crohn’s disease that may be bridged by the effects of treatment (Martin et al., 2019).

The availability of comprehensive clinical, flow cytometric and scRNA-seq data from patients with pediCD and from uninflamed FGID controls created an unprecedented opportunity for comparative atlas creation. We developed a methodical, unbiased, approach to cell state discovery, ARBOL, released alongside our manuscript in both R (https://github.com/jo-m-lab/ARBOL) and python (https://github.com/jo-m-lab/ARBOLpy). ARBOL iteratively explores axes of variation in scRNA-seq data by clustering and subclustering until specific stop conditions are met. The philosophy of ARBOL is that multiple axes of variation could be biologically meaningful at each tier, and that axes of variation are relative to the comparative outgroup, meaning that similar cell states may arise at distinct tiers. Once these possibilities are explored, curation and a statistical interrogation of resolution are used to collapse clusters into the elemental transcriptomes and co-varying gene expression of the dataset. ARBOL inherently builds a tree of subclustering events. As data is separated by major axes of variation in each subset, later rounds capture less pronounced variables. This comes with some caveats: variation shared by all cell types (for example, cell cycle stage) can make up one of the major axes of variation in the first round of clustering. Cell types can split up at the beginning, so the same splitting of B and T cells, for example, may happen further down in separate branches. The resulting tree of clustering events (Supplemental Figure 5a) is therefore neither indicative of true distances between end clusters nor a tree of unique groupings. We address this problem by calculation of a binary tree of manually and computationally curated end clusters. Using a standardized method of end cluster naming, which we describe in ARBOL’s tutorial (https://jo-m-lab.github.io/ARBOL/ARBOLtutorial.html), we found the resulting binary tree assorted end clusters into appreciated cell types and subsets (Figures 3 and 4), and also reveals further previously unappreciated granularity that will serve as the foundation for future work into the cellular composition of the gastrointestinal tract.

One of the primary remaining challenges going forward will be to identify which clusters are truly patient-unique, or are simply patient-unique at the cohort size to which we are currently limited to. We calculate a diversity metric for each end cluster to highlight those which are largely conserved between patients, and provide complete cluster-defining gene lists for both FGID and pediCD at three levels of clustering. We also provide links to our data visualization portal to enable cross-atlas comparisons: https://singlecell.broadinstitute.org/single_cell/study/SCP1422/predict-2021-paper-fgid and https://singlecell.broadinstitute.org/single_cell/study/SCP1423/predict-2021-paper-cd.

ARBOL can be deployed in the generation of cell atlases from multiple study designs to yield, for example, separate atlases for distinct diseases, joint atlases of diseases within a study, or integrated larger meta-atlases that merge multiple studies. Recent work from the integrated Human Lung Cell Atlas highlights that under-clustering of cells continues to be a large hurdle in many individual studies, but that higher clustering power must be balanced with the use of more stringent integration methods to merge control and disease groups without obscuring known subsets within lymphoid cells such as regulatory T cells and ILCs (Sikkema et al., 2022). One of the chief advantages of enrolling patients at diagnosis, and prior to any therapeutic intervention, was that we were able to relate their diagnostic immune landscape with disease trajectory. In the pediCD group, we identified 3 clinical subgroups. The first distinction was made by treating physicians, and classified patients with milder versus more severe clinical disease characteristics at diagnosis. The milder patients were not placed on anti-TNF agents (NOA), while the more severe patients were treated with monoclonal antibodies that neutralize TNF including infliximab and adalimumab. The second distinction between patient groups could not be made at diagnosis, but rather, was based on clinical and biochemical response to anti-TNF agents. Thus, of those patients treated with anti-TNF therapeutics, some were FRs, and some were only PRs, with PRs requiring anti-TNF dose modifications and the addition of other agents, and with ongoing, uncontrolled disease signs and symptoms. While differences in anti-TNF pharmacokinetics have been partially implicated in the need to dose-escalate anti-TNF agents in some pediCD patients, our study identifies foundational differences in the immune state at diagnosis in PR patients compared to the NOA and FR subgroups (D’Haens and Deventer, 2021; Ordás et al., 2012; Yarur et al., 2016). Although standard flow cytometry was not able to distinguish the immune phenotype of NOA versus treated patients, scRNA-seq identified significant differences. The contextualization of our scRNA-seq derived predictive cellular vector, pediCD-TIME, with two other treatment-naïve bulk RNA-seq studies of Crohn’s disease, and a reference adult scRNA-seq study, underscores the broader applicability of our findings (Kugathasan et al., 2017; Verstockt et al., 2019).

We noted significant cell state changes at diagnosis underlying clinically-appreciated disease severity that impacted the clinical decision to treat or not to treat with anti-TNF agents. These occurred within multiple clusters of T, NK, fibroblast, epithelial, monocyte, macrophage, and dendritic cells. For anti-TNF response, very few clusters exhibited significantly differential composition between FR and PR individuals. This suggests that multiple collective changes in several cell types may conspire to lead to differences in treatment outcomes. Indeed, when we jointly considered a cellular principal component vector comprising epithelial cells, T/NK/ILCs, and myeloid cells, we identified several clusters that together could delineate the full spectrum of NOA, FR, and PR. This cellular vector (pediCD-TIME) indicated that multiple T cell subsets, NK cells, monocytes, macrophages, and epithelial cells were altered in disease.

Epithelial cells involved in chemosensation (Tuft.GNAT3.TRPM5) and absorption of metabolites (EC.GSTA3.TMPRSS15), as well as stem cells (Banerjee et al., 2020; Sido et al., 1998; von Moltke et al., 2016) were enriched in NOA individuals. That pediCD severity is not uniquely predicted by a singular cell subset or gene is reflective of the complex genetics and environmental factors that have been implicated, along with the rich literature that has found significant changes by histology, flow cytometry, or mass cytometry in CD relative to control tissue (Buisine et al., 2001; Leeb et al., 2003; Leonard et al., 1995; Lilja et al., 2000; Mitsialis et al., 2020; Müller et al., 1998; Souza et al., 1999; Stappenbeck and McGovern, 2017; Takayama et al., 2010). However, with the PREDICT study, we have discovered precisely which changes in CD cellular composition collectively form a vector that anticipates both disease severity and treatment response. Intriguingly, the quantification and visualization of this response vector predicted a later escalation of one of our patients (p022; who appeared as an outlier FR in Figure 5d) from FR to PR, which occurred after our database lock. Furthermore, our approach also leverages co-variation of genes across single cells to identify gene sets which are not found by differential expression testing (which ignores associations between genes across cells) yet give rise to recurrent cell states found in multiple patients that are differentially abundant.

When considering the relationships between T cells and NK cells along with epithelial cells, we captured that proliferating cytotoxic NK cell subsets like CD.NK.MKI67.GZMA were significantly negatively correlated with critical metabolic and progenitor epithelial cell subsets in pediCD. Conversely, proliferating regulatory CD.T.MKI67.FOXP3 were positively associated with secretory epithelial cells in pediCD, but did not appear related to the decrease in metabolic or progenitor cells. In addition, we have found that pediCD samples have a significant increase in the number of CD4 T cells localized to the epithelial lining relative to FGID. How T cell-derived cytokines impact intestinal regeneration and differentiation has recently been the focus of several studies, but the relationship of these fine-grained T cell subsets with specific epithelial cell states observed in the human intestine remained unknown (Biton et al., 2018; Lindemans et al., 2015). Furthermore, our study builds on work from other groups who have recently started to describe specific features of intestinal intraepithelial T cells in inflammation and infection (Hoytema van Konijnenburg et al., 2017; Jaeger et al., 2021; Parsa et al., 2022). Our work suggests that there is further complexity to understand, particularly as it pertains to specific subsets of cytotoxic NK cells and T cells, and their impact on epithelial cell homeostasis and regeneration in pediCD.

The mapping of these disease severity-associated cell networks identifies a host of new potential therapeutic targets for pediCD, for many of which there are clinical-stage therapeutics that could be investigated. These include CD40L-blocking antibodies, IL-22 agonists, and targeted anti-proliferation agents (Betts et al., 2017; Furlan et al., 2015; Lindemans et al., 2015; Miura et al., 2021; Ramanujam et al., 2020; Sootome et al., 2020). A case can also be built for targeting inflammatory cytokines such as IL-1², and for interrogating agents aimed at mucosal healing including new anti-GM-CSF antibodies, given that several prominent cell subsets marked by CSF2 were enriched in the PR patients (Ai et al., 2021; Aschenbrenner et al., 2021; Castro-Dopico et al., 2020; Mehta et al., 2020; Mitsialis et al., 2020; Muro and Mrowiec, 2015). This atlas therefore provides a rigorous evidence-based rationale for proposing new therapeutic interventions, as well as a mechanism for interrogating the impact of new agents on the longitudinal immune landscape of pediCD patients. It also identifies those clusters which respond to anti-TNF treatment and may present new therapeutic targets.

Our analysis underscores the power in the disease-specific clustering employed with this dataset, which revealed principled end clusters in pediCD which could then be related back to a non-inflammatory reference atlas of FGID, and contextualized with published adult CD atlases; while still maintaining fine granularity through distinct computational methods. Recent work on COVID-19 has also highlighted the challenges faced by systems approaches to capture baseline cell states that predict disease trajectory (Kaczorowski et al., 2017; Lucas et al., 2020; Mathew et al., 2020; Schulte-Schrepping et al., 2020; Su et al., 2020). In a disease of known infectious etiology with SARS-CoV-2, monocytes, macrophages, granulocytes, T cells, B cells, antibodies, and interferon state have all independently been associated with disease outcomes. Enabled by larger numbers of participant biopsies profiled at single-cell resolution, recent work in cancer builds on analytical frameworks using dimensional reduction techniques on flow cytometry data to understand collective cellular changes in multi-cellular tissues (Combes et al., 2022; Kaczorowski et al., 2017; Nieto et al., 2021). This work has helped establish the paradigm of tissue “archetypes” or “immunotypes” that represent the collective cellular system. Here, we have built on this foundation to consider how multiple collective changes at baseline may influence outcome, yet are likely more reflective of the disease. With the complex and protracted presentation of a multifactorial disease like Crohn’s disease, our data suggest that multiple concerted effects are required to dictate both the severity (NOA vs FR/PR) and the treatment-response (FR vs PR). Future work will further consider which cell subsets are recovered during mucosal healing, and how closely the treated state reflects each individual patient’s baseline presentation. Our work taken together with our colleagues’ points to the emergent opportunity to place a patient’s tissue in the broader cellular context of health and disease.

Limitations of the Study

Our study has several limitations that are important to consider. Despite having profiled more treatment-naïve patients than any other studies at single-cell resolution, we are still limited by the cohort size to the strict inclusion/exclusion criteria used, follow-up period, and the ability to capture untreated patients early during the diagnosis process. As a result, we are underpowered to directly correlate in our cohort genetic and environmental influences such as the enteric microbiome, infections, and dietary factors (Dovrolis et al., 2020; Rajca et al., 2014; Yilmaz et al., 2019). Second, we are limited in our ability to capture full spatial transcriptomic information from our initial cohort. It will also be critical to understand how disease progresses in larger numbers of patients from our cohort, and the reason for persistent NOA, FR, or PR patient states in some individuals which will require retention in the trial and additional repeat biopsy procurement from larger numbers of participants. Third, our cohort is representative of the demographics of FGID and IBD in Seattle, WA. It will be essential to understand how both FGID and IBD may present differently across the world, and multiple cohorts will be required to understand location-unique and generalizable findings. Fourth, we also recognize the number of pediCD patients within this cohort limits our ability to subdivide CD phenotype into location and behavior of disease (via the Paris/Montreal classification systems) (Levine et al., 2011). Fifth, we also highlight the rapidly-evolving field of approaches to integrate single-cell datasets towards a larger gut cell atlas and the trade-offs in resolution that occur as additional datasets are brought in that require more stringent integration methods. We recognize that integration methods, as well as the use of UMAP to represent multiple PC’s in two dimensions, which, while helpful to jointly visualize studies together, may lower the resolution, and in some cases even make cell subsets like T, NK, or ILCs challenging to discern (Sikkema et al., 2022). Future work will need to understand specific platform differences that drive integration to help specifically separate batch from biology. Finally, as we are working with small pediatric IBD biopsies, our ability to functionally test and validate the stability of these end cell clusters was limited in this study. Based on our study, we will seek to expand our efforts to validate and extend our finding in ongoing work.

eLife Assessment

This important study substantially advances our understanding of pediatric Crohn's disease, mapping the cellular make-up of this disease and how patients respond to treatment. The evidence supporting the conclusions is compelling, with thorough bioinformatic analyses, underpinned by rigorous methodology and data integration. The work will be of broad interest to pediatric clinicians, immunologists and bioinformaticians.

Reviewer #1 (Public Review):

Summary: Crohn's disease is a prevalent inflammatory bowel disease that often results in patient relapse post anti-TNF blockades. This study employs a multifaceted approach utilizing single-cell RNA sequencing, flow cytometry, and histological analyses to elucidate the cellular alterations in pediatric Crohn's disease patients pre and post-anti-TNF treatment and comparing them with non-inflamed pediatric controls. Utilizing an innovative clustering approach, the research distinguishes distinct cellular states that signify the disease's progression and response to treatment. Notably, the study suggests that the anti-TNF treatment pushes pediatric patients towards a cellular state resembling adult patients with persistent relapses. This study's depth offers a nuanced understanding of cell states in CD progression that might forecast the disease trajectory and therapy response.

Robust Data Integration: The authors adeptly integrate diverse data types: scRNA-seq, histological images, flow cytometry, and clinical metadata, providing a holistic view of the disease mechanism and response to treatment.

Novel Clustering Approach: The introduction and utilization of ARBOL, a tiered clustering approach, enhances the granularity and reliability of cell type identification from scRNA-seq data.

Clinical Relevance: By associating scRNA-seq findings with clinical metadata, the study offers potentially significant insights into the trajectory of disease severity and anti-TNF response; which might help with the personalized treatment regimens.

Treatment Dynamics: The transition of the pediatric cellular ecosystem towards an adult, more treatment-refractory state upon anti-TNF treatment is a significant finding. It would be beneficial to probe deeper into the temporal dynamics and the mechanisms underlying this transition.

Comparative Analysis with Adult CD: The positioning of on-treatment biopsies between treatment-naïve pediCD and on-treatment adult CD is intriguing. A more in-depth exploration comparing pediatric and adult cellular ecosystems could provide valuable insights into disease evolution.

Areas of improvement:

1. The legends accompanying the figures are quite concise. It would be beneficial to provide a more detailed description within the legends, incorporating specifics about the experiments conducted and a clearer representation of the data points.

2. Statistical significance is missing from Fig. 1c WBC count plot, Fig. 2 b-e panels. Please provide it even if it's not significant. Also, the legend should have the details of stat test used.

3. In the study, the NOA group is characterized by patients who, after thorough clinical evaluations, were deemed to exhibit milder symptoms, negating the need for anti-TNF prescriptions. This mild nature could potentially align the NOA group closer to FIGD-a condition intrinsically defined by its low to non-inflammatory characteristics. Such an alignment sparks curiosity: is there a marked correlation between these two groups? A preliminary observation suggesting such a relationship can be spotted in Figure 6, particularly panels A and B. Given the prevalence of FIGD among the pediatric population, it might be prudent for the authors to delve deeper into this potential overlap, as insights gained from mild-CD cases could provide valuable information for managing FIGD.

4. Furthermore, Figure 7 employs multi-dimensional immunofluorescence to compare CD, encompassing all its subtypes, with FIGD. If the data permits, subdividing CD into PR, FR, and NOA for this comparison could offer a more nuanced understanding of the disease spectrum. Such a granular perspective is invaluable for clinical assessments. The key question then remains: do the sample categorizations for the immunofluorescence study accommodate this proposed stratification?

5. The study's most captivating revelation is the proximity of anti-TNF-treated pediatric CD (pediCD) biopsies to adult treatment-refractory CD. Such an observation naturally raises the question: How does this alignment compare to a standard adult colon, and what proportion of this similarity is genuinely disease-specific versus reflective of an adult state? To what degree does the similarity highlight disease-specific traits?

Delving deeper, it will be of interest to see whether anti-TNF treatment is nudging the transcriptional state of the cells towards a more mature adult stage or veering them into a treatment-resistant trajectory. If anti-TNF therapy is indeed steering cells toward a more adult-like state, it might signify a natural maturation process; however, if it's directing them toward a treatment-refractory state, the long-term therapeutic strategies for pediatric patients might need reconsideration.

Reviewer #2 (Public Review):

Summary:

Through this study, the authors combine a number of innovative technologies including scRNAseq to provide insight into Crohn's disease. Importantly samples from pediatric patients are included. The authors develop a principled and unbiased tiered clustering approach, termed ARBOL. Through high-resolution scRNAseq analysis the authors identify differences in cell subsets and states during pediCD relative to FGID. The authors provide histology data demonstrating T cell localisation within the epithelium. Importantly, the authors find anti-TNF treatment pushes the pediatric cellular ecosystem toward an adult state.

Strengths:

This study is well presented. The introduction clearly explains the important knowledge gaps in the field, the importance of this research, the samples that are used, and study design.

The results clearly explain the data, without overstating any findings. The data is well presented. The discussion expands on key findings and any limitations to the study are clearly explained.

I think the biological findings from, and bioinformatic approach used in this study, will be of interest to many and significantly add to the field.

Weaknesses:

1. The ARBOL approach for iterative tiered clustering on a specific disease condition was demonstrated to work very well on the datasets generated in this study where there were no obvious batch effects across patients. What if strong batch effects are present across donors where PCA fails to mitigate such effects? Are there any batch correction tools implemented in ARBOL for such cases?

2. The authors mentioned that the clustering tree from the recursive sub-clustering contained too much noise, and they therefore used another approach to build a hierarchical clustering tree for the bottom-level clusters based on unified gene space. But in general, how consistent are these two trees?

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

Crohn's disease is a prevalent inflammatory bowel disease that often results in patient relapse post anti-TNF blockades. This study employs a multifaceted approach utilizing single-cell RNA sequencing, flow cytometry, and histological analyses to elucidate the cellular alterations in pediatric Crohn's disease patients pre and post-anti-TNF treatment and comparing them with non-inflamed pediatric controls. Utilizing an innovative clustering approach, the research distinguishes distinct cellular states that signify the disease's progression and response to treatment. Notably, the study suggests that the anti-TNF treatment pushes pediatric patients towards a cellular state resembling adult patients with persistent relapses. This study's depth offers a nuanced understanding of cell states in CD progression that might forecast the disease trajectory and therapy response.

Robust Data Integration: The authors adeptly integrate diverse data types: scRNA-seq, histological images, flow cytometry, and clinical metadata, providing a holistic view of the disease mechanism and response to treatment.

Novel Clustering Approach: The introduction and utilization of ARBOL, a tiered clustering approach, enhances the granularity and reliability of cell type identification from scRNA-seq data.

Clinical Relevance: By associating scRNA-seq findings with clinical metadata, the study offers potentially significant insights into the trajectory of disease severity and anti-TNF response; which might help with the personalized treatment regimens.

Treatment Dynamics: The transition of the pediatric cellular ecosystem towards an adult, more treatment-refractory state upon anti-TNF treatment is a significant finding. It would be beneficial to probe deeper into the temporal dynamics and the mechanisms underlying this transition.

Comparative Analysis with Adult CD: The positioning of on-treatment biopsies between treatment-naïve pediCD and on-treatment adult CD is intriguing. A more in-depth exploration comparing pediatric and adult cellular ecosystems could provide valuable insights into disease evolution.

Areas of improvement:

(1) The legends accompanying the figures are quite concise. It would be beneficial to provide a more detailed description within the legends, incorporating specifics about the experiments conducted and a clearer representation of the data points.

We agree that it is beneficial to have descriptive figure legends that balance elements of experimental design, methodology, and statistical analyses employed in order to have a clear understanding throughout the manuscript. We have gone through and clarified areas throughout.

(2) Statistical significance is missing from Fig. 1c WBC count plot, Fig. 2 b-e panels. Please provide it even if it's not significant. Also, the legend should have the details of stat test used.

We have now added details of statistical significance data in the Figure 1 legends. Please note that Mann-Whitney U-test was used for clinical categorical data.

(3) In the study, the NOA group is characterized by patients who, after thorough clinical evaluations, were deemed to exhibit milder symptoms, negating the need for anti-TNF prescriptions. This mild nature could potentially align the NOA group closer to FGID-a condition intrinsically defined by its low to non-inflammatory characteristics. Such an alignment sparks curiosity: is there a marked correlation between these two groups? A preliminary observation suggesting such a relationship can be spotted in Figure 6, particularly panels A and B. Given the prevalence of FGID among the pediatric population, it might be prudent for the authors to delve deeper into this potential overlap, as insights gained from mild-CD cases could provide valuable information for managing FGID.

Thank you for this insightful point. On histopathology and endoscopy, the NOA exhibited microscopic and macroscopic inflammation which landed these patients with the CD diagnosis, albeit mild on both micro and macro accounts. By contrast, the FGID group by definition will not have inflammation of microscopic and macroscopic evaluation. There is great interest in the field of adult and pediatric gastroenterology to understand why patients develop symptoms without evidence of inflammation. However, in 2023 the diagnostic tools of endoscopy with biopsy and histopathology is not sensitive enough to detect transcript level inflammation, positioning single-cell technology to be able to reveal further information in both disease processes.

Based on the reviewer’s suggestions, we have calculated a heatmap of overlapping NOA and FGID cell states along the Figure 6a joint-PC1, showing where NOA CD patients and FGID patients overlap in terms of cell states. This is displayed in Supplemental Figure 15d. This revealed a set of T, Myeloid, and Epithelial cell states that were most important in describing variance along the FGID-CD axis, allowing us to hone in on similarities at the boundary between FGID and CD. By comparing the joint cell states with CD atlas curated cluster names, we identified CCR7-expressing T cell states and GSTA2-expressing epithelial states associated with this overlap.

(4) Furthermore, Figure 7 employs multi-dimensional immunofluorescence to compare CD, encompassing all its subtypes, with FGID. If the data permits, subdividing CD into PR, FR, and NOA for this comparison could offer a more nuanced understanding of the disease spectrum. Such a granular perspective is invaluable for clinical assessments. The key question then remains: do the sample categorizations for the immunofluorescence study accommodate this proposed stratification?

Thank you for the thoughtful discussion. We agree that stratifying Crohn’s disease by PR, FR, and NOA would provide valuable clinical insight. Unfortunately our multiplex IF cohort was designed to maximize overall CD versus FGID comparisons and does not contain enough samples in patient subgroups to power such an analysis. We have highlighted this limitation in the text.

(5)The study's most captivating revelation is the proximity of anti-TNF-treated pediatric CD (pediCD) biopsies to adult treatment-refractory CD. Such an observation naturally raises the question: How does this alignment compare to a

standard adult colon, and what proportion of this similarity is genuinely disease-specific versus reflective of an adult state? To what degree does the similarity highlight disease-specific traits?

Delving deeper, it will be of interest to see whether anti-TNF treatment is nudging the transcriptional state of the cells towards a more mature adult stage or veering them into a treatment-resistant trajectory. If anti-TNF therapy is indeed steering cells toward a more adult-like state, it might signify a natural maturation process; however, if it's directing them toward a treatment-refractory state, the long-term therapeutic strategies for pediatric patients might need reconsideration.

Thank you to the reviewer for another insightful point. We agree that age-matched samples are critical to evaluate disease cell states and hence we have age-matched controls in our pediatric cohort. Our timeline of follow-up only spans 3 years and patients remain in the pediatric age range at times of follow-up endoscopy and biopsy and would not be reflective of an adult GI state. We believe that the cellular behavior from naïve to treatment biopsy to on treatment biopsy is reflective of disease state rather than movement towards and adult-like state. We would also like to point out that pediatric onset IBD (Crohn’s and ulcerative colitis) traditionally has been harder to treat and presents with more extensive disease state (PMID: 22643596) and the ability to detect need for therapy escalation/change would be an invaluable tool for clinicians.

We share the reviewer’s interest in disentangling a natural maturation process from disease and treatment-specific changes. Because the patients who were not given treatment did not move towards the adult-like phenotype, it could point to a push towards a treatment-resistant trajectory. To further support these findings, we generated a new disease-pseudotime figure Supplemental Figure 17, using cross-validation methods and the TradeSeq package. This figure was designed to track how each pediatric sample shifts from the treatment-naïve state through antiTNF therapy and to test the robustness of these shifts across samples. The new visualizations show patterns that do not recapitulate natural aging processes but rather shifts across all cell types associated with antiTNF treatment.

Reviewer #2 (Public Review):

Summary:

Through this study, the authors combine a number of innovative technologies including scRNAseq to provide insight into Crohn's disease. Importantly samples from pediatric patients are included. The authors develop a principled and unbiased tiered clustering approach, termed ARBOL. Through high-resolution scRNAseq analysis the authors identify differences in cell subsets and states during pediCD relative to FGID. The authors provide histology data demonstrating T cell localisation within the epithelium. Importantly, the authors find anti-TNF treatment pushes the pediatric cellular ecosystem toward an adult state.

Strengths:

This study is well presented. The introduction clearly explains the important knowledge gaps in the field, the importance of this research, the samples that are used, and study design.

The results clearly explain the data, without overstating any findings. The data is well presented. The discussion expands on key findings and any limitations to the study are clearly explained.

I think the biological findings from, and bioinformatic approach used in this study, will be of interest to many and significantly add to the field.

Weaknesses:

(1) The ARBOL approach for iterative tiered clustering on a specific disease condition was demonstrated to work very well on the datasets generated in this study where there were no obvious batch effects across patients. What if strong batch effects are present across donors where PCA fails to mitigate such effects? Are there any batch correction tools implemented in ARBOL for such cases?

We thank the reviewer for their insightful point, the full extent to which ARBOL can address batch effects requires further study. To this end we integrated Harmony into the ARBOL architecture and used it in the paper to integrate a previous study with the data presented (Figure 8). We have added to ARBOL’s github README how to use Harmony with the automated clustering method. With ARBOL, as well as traditional clustering methods, batch effects can cause artifactual clustering at any tier of clustering. Due to iteration, this can cause batch effects to present themselves in a single round of clustering, followed by further rounds of clustering that appear highly similar within each batch subset. Harmony addresses this issue, removing these batch-related clustering rounds. The later arrangement of fine-grained clusters using the bottom-up approach can use the batch-corrected latent space to calculate relationships between cell states, removing the effects from both sides of the algorithm. As stated, the extent to which ARBOL can be used to systematically address these batch effects requires further research, but the algorithmic architecture of ARBOL is well suited to address these effects.

(2) The authors mentioned that the clustering tree from the recursive sub-clustering contained too much noise, and they therefore used another approach to build a hierarchical clustering tree for the bottom-level clusters based on unified gene space. But in general, how consistent are these two trees?

Thank you for this thoughtful question. The two tree methodologies are not consistent due to their algorithmic differences, but both are important for several reasons:

(1) The clustering tree is top-down, meaning low resolution lineage-related clusters are calculated first. Doublets and quality differences can cause very small clusters of different lineages (endothelial vs fibroblast) to fall under the incorrect lineage at first in the sub clustering tree, but these are recaptured during further sub clustering rounds, and then disentangled by the cluster-centroid tree.

(2) The hierarchical tree is a rose tree, meaning each branching point can contain several daughter branches, while taxonomies based on distances between species (or cell types in this case) are binary trees with only 2 branches per branching point, because distances between each cluster are unique. Because this taxonomy, or bottom-up, is different from the top-down approach, it is useful to then look at how these bottom-level clusters are similar. To that end, we performed pair-wise differential expression between all end clusters and clustered based on those genes.

(3) Calculation of a binary tree represents a quantitative basis for comparing the transcriptomic distance between clusters as opposed to relying on distances calculated within a heuristic manifold such as UMAP or algorithmic similarity space such as cluster definitions based on KNN graphs.

In practice, this dual view rescues small clusters that may have been mis-grouped by technical artifacts and gives a quantitative distance based hierarchy that can be compared across metadata covariates.

Version 2

Abstract

Crohn’s disease is an inflammatory bowel disease (IBD) commonly treated through anti-TNF blockade. However, most patients still relapse and inevitably progress. Comprehensive single-cell RNA-sequencing (scRNA-seq) atlases have largely sampled patients with established treatment-refractory IBD, limiting our understanding of which cell types, subsets, and states at diagnosis anticipate disease severity and response to treatment. Here, through combining clinical, flow cytometry, histology, and scRNA-seq methods, we profile diagnostic human biopsies from the terminal ileum of treatment-naïve pediatric patients with Crohn’s disease (pediCD; n=14), matched repeat biopsies (pediCD-treated; n=8) and from non-inflamed pediatric controls with functional gastrointestinal disorders (FGID; n=13). To resolve and annotate epithelial, stromal, and immune cell states among the 201,883 baseline single-cell transcriptomes, we develop a principled and unbiased tiered clustering approach, ARBOL. Through flow cytometry and scRNA-seq, we observe that treatment-naïve pediCD and FGID have similar broad cell type composition. However, through high-resolution scRNA-seq analysis and microscopy, we identify significant differences in cell subsets and states that arise during pediCD relative to FGID. By closely linking our scRNA-seq analysis with clinical meta-data, we resolve a vector of T cell, innate lymphocyte, myeloid, and epithelial cell states in treatment-naïve pediCD (pediCD-TIME) samples which can distinguish patients along the trajectory of disease severity and anti-TNF response. By using ARBOL with integration, we position repeat on-treatment biopsies from our patients between treatment-naïve pediCD and on-treatment adult CD. We identify that anti-TNF treatment pushes the pediatric cellular ecosystem towards an adult, more treatment-refractory state. Our study jointly leverages a treatment-naïve cohort, high-resolution principled scRNA-seq data analysis, and clinical outcomes to understand which baseline cell states may predict Crohn’s disease trajectory.

Main

Inflammatory bowel diseases (IBDs) arise when homeostatic mechanisms regulating gastrointestinal (GI) tract tissue integrity, nutrient absorption, and protective immunity are replaced by pathogenic inflammation (Baumgart and Sandborn, 2012; Chang, 2020; Corridoni et al., 2020b; Friedrich et al., 2019; Graham and Xavier, 2020; Selin et al., 2021). The initiating triggers are not fully known, but host genetics and the microbiome are being increasingly appreciated to play important, and in some cases causal roles in the IBDs (Chang, 2020; Cohen et al., 2019; Franzosa et al., 2019; Jain et al., 2021; Limon et al., 2019). Crohn’s disease (CD) presents predominantly in the terminal ileum and the proximal colon, though lesions may develop anywhere along the gastrointestinal tract (Baumgart and Sandborn, 2012; Chang, 2020; Kobayashi et al., 2020; Roda et al., 2020). Among the IBDs, pediatric-onset Crohn’s disease (pediCD) is particularly common (25% of all IBD cases, 60-70% of pediatric IBD) and is a debilitating form due to its early presentation, impact on the terminal ileum and proximal colon, and the lack of disease-specific therapies developed for children (Hyams et al., 1991; Ruemmele et al., 2014; Sýkora et al., 2018; Turner et al., 2012; Ye et al., 2020). In contrast to pediCD, there exists a group of pediatric disorders termed functional gastrointestinal disorders (FGIDs), which include GI symptoms that necessitate endoscopy, but lack laboratory markers, endoscopic findings, and histologic evidence associated with inflammation (Black et al., 2020; Hyams et al., 2016; McOmber and Shulman, 2008; Santucci et al., 2020). Thus, while FGID is a distinct disease entity that does not represent completely healthy tissue, it is a critical non-inflamed age-matched control cohort with which to contextualize the inflammation observed in pediCD.

The current standard of care for pediCD (as with adult CD) is tailored to the patient’s disease location, clinical behavior and severity, though use of prednisone and other immunomodulators, as well as monoclonal antibodies including anti-TNFα, are common (Hyams et al., 1991; Ruemmele et al., 2014; Turner et al., 2012). While targeting TNF is shared across many autoimmune and inflammatory conditions, it is not successful in all patients, and many go on to develop anti-TNF-refractory disease. It is of tremendous importance for the field to precisely understand and characterize for which patients anti-TNF therapy is not necessary, in which it may succeed in controlling disease, and which will likely be refractory to treatment. Several ideas based on individual immunogenicity and pharmacokinetics have been proposed to explain TNF-refractory disease, including gender (M>F), low albumin levels, high BMI, and high baseline C-Reactive Protein (CRP) (Atreya et al., 2020; Digby-Bell et al., 2020). However, no single identifiable clinical or biochemical biomarker reliably predicts disease response versus resistance to anti-TNF treatment (Stevens et al., 2018).

The primary cellular lineages sampled from intestinal biopsies of CD patients represent both the epithelium and lamina propria, and include epithelial cells, stromal cells, hematopoietic cells and neuronal processes whose cell bodies are present outside of these regions (Buisine et al., 2001; Leeb et al., 2003; Leonard et al., 1995; Lilja et al., 2000; Müller et al., 1998; Souza et al., 1999; Stappenbeck and McGovern, 2017; Takayama et al., 2010). Alterations to all cellular lineages have been implicated in CD (Furey et al., 2019; Martin et al., 2019). Histologically, CD is characterized by a granulomatous inflammation, and by alterations in almost every leukocyte cell type studied, including an increase of cytotoxic lymphocytes, alterations in γδ T cells, increases in mast cells and their production of TNF, activation and shifts in antibody isotypes towards IgM and IgG from B cells and plasma cells, and cytokine production by macrophages (Catalan-Serra et al., 2017; Lilja et al., 2000; Meijer et al., 1979; Mitsialis et al., 2020; Müller et al., 1998; Sieber et al., 1984; Takayama et al., 2010). In the stromal compartment, there is evidence for enhanced vascularization and increased expression of ICAM-1 and MAdCAM-1 by vascular beds, substantial remodeling of collecting lymphatics, and altered migratory potential of fibroblasts (Leeb et al., 2003; Souza et al., 1999). Epithelial barrier dysfunction has also been noted, including alterations to mucus production, microvilli, and Paneth cell dysfunction (Buisine et al., 2001; Stappenbeck and McGovern, 2017). Previous studies have therefore identified that essentially all cell types may be meaningfully altered during CD, highlighting the important need to comprehensively chart the concerted cellular changes that define CD, its severity, and its expected treatment course. Importantly, which changes are predictive of more severe disease, response to treatment, or treatment resistance in pediCD remain largely unknown.

Single-cell RNA-sequencing (scRNA-seq) is enhancing our ability to comprehensively map and resolve the cell types, subsets, and states present during health and disease. Recent work has generated cellular atlases of adult CD and UC from patients who were already on treatment and without linking to clinical follow-up studies (Corridoni et al., 2020b, 2020a; Drokhlyansky et al., 2019; Elmentaite et al., 2020; Huang et al., 2019; Kinchen et al., 2018; Martin et al., 2019; Parikh et al., 2019; Smillie et al., 2019). The potential impact of scRNA-seq on our understanding of IBD is evidenced by studies of adult UC, which have identified potential functional roles for poorly understood colonic cell subsets, such as the BEST4+ enterocyte, and identified pathological alterations in UC biopsies compared to healthy controls, including an expansion of microfold-like cells, IL13RA2+IL11+ inflammatory fibroblasts, CD4+CD8+IL17A+ T cells and CD8+GZMK+ T cells (Smillie et al., 2019). A single-cell study of on-treatment adult CD patients comparing non-inflamed and inflamed tissue from surgically-resected bowel identified IgG+ plasma cells, inflammatory mononuclear phagocytes, activated T cells and stromal cells, comprising the “GIMATS” pathogenic cellular module (Martin et al., 2019). This module was used to derive a gene signature associated with resistance to anti-TNF therapy in a distinct cohort profiled by bulk RNA-seq. Very recent work has profiled how fetal transcription factors are reactivated in Crohn’s disease epithelium, leading to significant decreases in mature absorptive epithelial cells and increases in secretory/goblet cells (Elmentaite et al., 2020). Together, these and other studies have demonstrated the power of scRNA-seq to nominate individual and collective cell states that are associated with disease, and underscored the unmet need to apply these techniques to untreated disease and associate them with disease severity in order to more specifically identify pathognomonic and prognostic cell states.

Most comprehensive scRNA-seq atlases of inflammatory disease conditions include patients being treated with a variety of agents, and for which the biopsies often reflect a partial treatment-refractory state to combinations of antibiotics, corticosteroids, immunomodulators, and biologics including anti-TNF monoclonal antibodies. A treatment-naïve single-cell atlas in an inflammatory disease condition linking observed baseline cell clusters with disease trajectory and treatment outcomes has yet to be reported. In order to address this unmet need in pediCD, we created the prospective PREDICT study (Clinicaltrials.gov #NCT03369353) to identify, profile, and understand pediatric IBD, as well as uninflamed FGID controls. Here, we present detailed diagnostic data from the first cohort of 27 patients enrolled on PREDICT, including 14 pediCD and 13 FGID patients, together with flow cytometric and scRNA-seq studies of the cellular composition of the terminal ileum (Figure 1). Furthermore, through prospective annotation of clinical metadata and detailed longitudinal follow-up, we stratify our pediCD cohort by clinically-guided therapeutic decisions separating patients treated with anti-TNF mAbs versus those with biopsy-proven pediCD, but for whom clinical symptoms were sufficiently mild that the treating physician did not prescribe anti-TNF agents (this cohort is termed “Not On Anti-TNF” or “NOA”). Importantly, we were also able to separate the cohort of patients treated with anti-TNF agents into sub-cohorts of those who achieved a full response (FR) to this therapy versus those who achieved only a partial response (PR). Because PREDICT enrolled patients prior to their diagnostic endoscopy, we were able to relate these clinical outcomes to the patients’ cell states at diagnosis, and were further able to integrate matched on-treatment biopsies from these same pediCD patients. We contextualize our findings in pediCD relative to a cohort of FGID patients, which provides an age-matched comparator cohort with clinical GI symptoms, but non-inflammatory disease proven by endoscopy and histologic examination.

PREDICT Study Design with Patient Diagnostic Criteria and Histopathology.

a. Study overview depicting clinical and cellular measurements from 13 functional gastrointestinal disorder (FGID) patients and 14 pediatric Crohn’s disease (pediCD) patients. Terminal ileum biopsies were isolated at a treatment-naïve diagnostic visit, and pediCD patients were followed up to determine their anti-TNF response and categorized as not on anti-TNF (NOA), Full Response (FR), or Partial Response (PR) (see Methods). Two panels of flow cytometry allowed for relative frequency quantification of 32 cell types and subsets, and 10X 3’ v2 single-cell RNA-sequencing (scRNA-seq) captured 254,911 total cells including 118 FGID and 305 pediCD end clusters through ARBOL (see Methods). b. Demographic data, weight, height, and BMI for cohort (see Table 1 and Supplemental Figure 1); Significance testing by Mann-Whitney U-test. No significance noted in age, weight, height, or BMI z-score for the two cohorts. c. Clinical inflammatory laboratory values for cohort (see Table 1 and Supplemental Figure 1); Significance testing by Mann-Whitney U-test. * indicates p<0.05. ** indicates p <0.005. Significance noted for C-reactive protein (CRP), hemoglobin (Hgb), end sedimentation rate (ESR), albumin, and platelets. No significance noted for WBC. d. Representative histopathology of FGID (top) and pediCD (bottom) at 10x (scale bar = 100um) and 40x (scale bar = 20um) magnification. e. Representative treatment history and clinical inflammatory parameters used for determination of NOA (p027), FR (p011) and PR (p018) status (see Methods, Table 1, and Supplemental Figure 1; ADA: adalimumab, INF: infliximab; MTX: methotrexate; Pred: prednisone; mSCD: modified specific carbohydrate diet; EEN: exclusive enteral nutrition).

Several analytical approaches have been developed to enable the generation and interrogation of clusters during the curation of single-cell transcriptomic atlases (Hie et al., 2020). One such method, sub-clustering of broad clusters, has proven to be a powerful tool for isolating highly specific axes of variation that are obscured by analyses whose principal axes of variation are broad cell types (Bakken et al., 2021; La Manno et al., 2021; Sikkema et al., 2022; Tasic et al., 2018; Zeisel et al., 2018). However, while sub-clustering analysis is a powerful tool allowing access to the hierarchy of cell states, this method is manually intensive and there is little consensus or standardization in clustering parameters and annotation methods. To address this issue, we designed a principled, modular, automated, iterative sub-clustering routine made possible by application of parameter scanning methods (Rousseeuw, 1987; Shekhar et al., 2016). We developed this tool, ‘ARBOL’ (named after tree in Spanish), of which iterative tiered clustering (ITC) is a key component, in R and python, integrating with Seurat and Scanpy functions, to make it accessible and easily incorporated into common workflows, and have curated a GitHub repository with illustrative vignettes. Here, we use ARBOL to standardize fine-grained cell state discovery by the creation and cultivation of a tree of cell states, followed by the automated generation of cell names to aid in the annotation of end clusters by unique and descriptive genes, inclusive of a method for identifying clusters containing high or low patient diversity, and for pruning of the resultant tree.

In order to respect the distinct clinical entities of FGID and pediCD, and the co-variation structure that can define rare but important cell subsets and states, we generate two fully-annotated cellular atlases for these pediatric GI diseases. These atlases consist of 94,451 cells for FGID and 107,432 for pediCD. We provide key gene-list resources for further studies, and nominate a vector of lymphoid, myeloid and epithelial cell states that anticipate disease severity and treatment outcomes. We term this vector ‘pediCD T cell, innate lymphocyte, myeloid and epithelial’ (pediCD-TIME). This cellular vector correlates strongly with both the clinical presentation of pediCD severity, and with the distinction between anti-TNF full-or partial-response. The significant changes in cell composition we have discovered to be associated with disease severity are increases of proliferating T cells, cytotoxic NK cells, subsets of monocytes/macrophages, and plasmacytoid dendritic cells (pDCs), accompanied by decreases of metabolically-specialized epithelial cell subsets. We successfully validate this vector in two bulk RNA-seq treatment-naïve IBD cohorts. We contextualize our treatment-naïve pediCD atlas by joining it with the FGID atlas through label-retaining (manual clustering; Random Forest; ARBOL) and label-free (Hellinger distance on UMAP) methods. Finally, we integrate our pediCD atlas with follow-up post-treatment biopsies from our cohort, and with the adult on-treatment CD patients of the “GIMATS” scRNA-seq atlas. Taken together, we identify that anti-TNF treatment over time pushes the overall pediCD cellular ecosystem towards the more treatment-refractory disease state found in adult CD.

Results

Study cohort clinical outcomes

The PREDICT study prospectively enrolled treatment-naïve, previously undiagnosed pediatric patients with GI symptoms necessitating diagnostic endoscopy. The current analysis focuses on patients enrolled in the first year of the study, during which time 14 patients with pediCD and 13 patients with FGID were enrolled and had adequate ileal samples for single-cell analysis (Figure 1; Supplemental Figure 1). After their initial diagnosis, patients with pediCD were followed clinically for up to 3 years (with the determination of anti-TNF response made at 2 years). Patients with FGID were followed clinically only as needed. The median time from diagnosis for the pediCD and FGID cohorts to the time of database lock was 32.5 and 31 months, respectively. Of the pediCD and FGID patients analyzed, the median age at diagnosis was 12.5 years and 16 years respectively (p = 0.095; Mann-Whitney U-test), with no significant differences in gender (Figure 1b; Table 1). Patient weight, height, and BMI z-scores were not significantly different between pediCD and FGID (Figure 1b; Table 1); however, in addition to the diagnostic differences on histologic analysis, several key clinical laboratory values, including C-reactive protein (CRP), Erythrocyte Sedimentation Rate (ESR), hemoglobin concentration, albumin concentration, and platelet count were significantly different between pediCD and FGID (Figure 1c,d; Table 1).

Treatment with anti-TNF agents and response to therapy

Patients with pediCD were initially divided into two cohorts. Those with milder disease characteristics (n = 4) as determined by their treating physician, were not treated with anti-TNF therapies, and are noted as ‘NOA’. For patients with more severe disease (n = 10), anti-TNF therapy (with either infliximab or adalimumab, Table 2) was initiated within 90 days of diagnostic endoscopy. All pediCD patients were followed prospectively and categorized as FR (n = 5) or PR (n = 5) to anti-TNF therapy based on the following criteria: FR was defined as clinical symptom control and biochemical response (measuring CRP, ESR, albumin, and complete blood counts (CBC)), and with a weighted Pediatric Crohn’s Disease Activity Index (PCDAI) score of <12.5 on maintenance anti-TNF therapy with no dose adjustments required (Cappello and Morreale, 2016; Hyams et al., 1991; Sandborn, 2014; Turner et al., 2017, 2012). PR to anti-TNF therapy was defined as a lack of full clinical symptom control as determined by the treating physician or lack of full biochemical response, with documented escalation of anti-TNF therapy or addition of other agents (Figure 1e; NB: patients in our cohort were dose-escalated because of clinical symptoms). Medication timelines and clinical laboratory data through 2 years post-diagnosis for all pediCD patients is shown in Supplemental Figure 1. The designation of FR or PR was made at 2 years of follow-up for all pediCD patients. During our study window, 8 pediCD (2 NOA, 3 FR, 3 PR) received repeat biopsies which we also analyzed through scRNA-seq (see below).

Flow cytometry of the terminal ileum reveals minimal changes in leukocyte subsets in FGID vs. pediCD, and no significant differences across the pediCD spectrum

We collected terminal ileum biopsies from 14 pediCD patients and 13 uninflamed FGID patients, and prepared single-cell suspensions for flow cytometry and scRNA-seq. Biopsies from pediCD were from actively-inflamed areas adjacent to ulcerations. Biopsies from FGID were from non-inflamed terminal ileum. The epithelium was first separated from the lamina propria before enzymatic dissociation, and flow cytometric analysis was performed on the viable single-cell fraction, which recovered predominantly hematopoietic cells with some remnant epithelial cells (<20% of all cells), likely representing those in deeper crypt regions (Figure 2; Supplemental Figure 2). We utilized two flow cytometry panels, allowing us to resolve the principal lymphoid (CD4 or CD8 T cells, NK cells, B cells, innate lymphoid cells, γδ T cells, CD8αα+ IELs, pDCs) and myeloid (monocytes, granulocytes, HLA-DR+ mononuclear phagocyte) cell subsets (Supplemental Figure 2, Supplemental Table 1). From these panels, which generated 32 gates identifying cell lineages, types and subsets, only HLA-DR+ macrophages/DCs and pDCs were significantly increased in pediCD relative to FGID (Figure 2d). We also analyzed within pediCD, comparing the baseline samples of 4 NOA, 5 FR and 5 PR patients, and noted no significant differences between NOA and patients on anti-TNF, or between FRs and PRs to anti-TNF (Figure 2e). Together, this suggests that despite the substantial endoscopic, histologic, and clinical parameters that distinguish FGID and pediCD (Figure 1), the broad single-cell type composition of the terminal ileum appears minimally altered in pediCD, save for an increased frequency of pDCs and HLA-DR+ mononuclear phagocytes.

Flow Cytometry of Ileal Biopsies Does Not Reveal Significant Changes in Cell Composition in FGID vs. pediCD or across the pediCD Treatment Response Spectrum.

a. Representative flow cytometry end gates for selected cell subsets (left: epithelial and hematopoietic; middle: naïve and effector T cells; right: pDCs and antigen-presenting cells) from single-cell dissociated samples from one terminal ileum biopsy for pediCD patients (see Supplemental Figure 2 for full gating strategy; numbers represent percentage of cells in each gate). b. Fractional composition of selected cell subsets of CD45+ cells from 13 FGID and 14 pediCD patients (error bars are s.e.m). c. Fractional composition of selected cell subsets of CD45+ cells from 4 NOA, 5 FR and 5 PR patients. d. Fractional composition of dendritic, pDC, central memory (CM) and effector memory (EM) CD4+ and CD8+ cells from 13 FGID vs 14 pediCD patients. Dendritic cells and pDC plotted as percentage of CD45+ cells. CM/EM CD4+ and CD8+ cells plotted as percentage of total CD4+ and CD8+ cells, respectively. p < 0.05 by Mann-Whitney for pediCD versus FGID and 1-way ANOVA for pediCD cohorts). e. Fractional composition of dendritic cells, pDCs, central memory (CM) and effector memory (EM) CD4+ and CD8+ cells from 4 NOA, 5FR, and 5 PR patients. Graphs plotted as in d.

Traditional clustering of scRNA-seq data from FGID and pediCD patients

In addition to flow cytometry, we performed droplet-based scRNA-seq on cell suspensions from the 14 pediCD and 13 FGID patients using the 10X Genomics 3’ V2 platform (Figure 1). The analyzed cell suspensions were derived from lamina propria preparations, which our flow cytometry data suggested would be composed primarily of CD45+ leukocytes, alongside a fraction of epithelial cells and stromal/vascular cells. Deconstructing these tissues into their component cells provided us with the ability to identify some of the corresponding cell types (e.g. T or B cell) and subsets (CD8αα+ IEL or CD4+ T cell) to those we identified by flow cytometry. Importantly, it also enabled us to: 1. characterize these major cell types and subsets without needing to pre-select markers, and 2. gain substantially enhanced resolution into the cell states (i.e. gene expression programs determined by co-expression of genes) within these cell types and subsets.

Following library preparation and sequencing, we derived a unified cells-by-genes expression matrix from the 27 samples, containing digital gene expression values for all cells passing quality thresholds (n=254,911 cells; Supplemental Figure 3; Supplemental Tables 2 and 3; Methods). We then performed dimensionality reduction and graph-based clustering, noting that despite no computational integration methods being used, FGID and pediCD were highly similar to each other when clustered together and visualized on a uniform manifold approximation and projection (UMAP) plot (Supplemental Figure 4a-c). We recovered the following cell types from both patient groups: epithelial cells, T cells, B cells, plasma cells, glial cells, endothelial cells, myeloid cells, mast cells, fibroblasts, and a proliferating cell cluster. We noted that the fractional composition amongst all cells of T cells, B cells, and myeloid cells was not significantly different between FGID and pediCD, similar to the flow cytometric data, and this was also the case for endothelial, fibroblasts, glial, mast and plasma cells, which were not measured through flow cytometry (Supplemental Figure 4d). This provided validation and extension of our flow cytometry data documenting that the broad cell type composition of FGID and pediCD is not significantly altered, despite highly-distinct endoscopic and histologic diseases. Based on this joint clustering and annotation of top-level cell types, we then performed differential expression testing identifying significant up- and down-regulated genes across cell types (Supplemental Figure 4e; Supplemental Table 4). Within myeloid cells we identified some of the most significantly upregulated genes in pediCD versus FGID to be CXCL9 and CXCL10, canonical IFNγ-stimulated genes, and S100A8 and S100A9 which form the biomarker fecal calprotectin (Supplemental Figure 4f)(Leach et al., 2007; Ziegler et al., 2021). Within epithelial cells we identified that APOA1 and APOA4 were amongst the most significantly downregulated genes, and correspondingly REG1B, SPINK4 and REG4A were amongst the most significantly upregulated indicating tradeoffs between lipid metabolism and host defense in pediCD versus FGID (Supplemental Figure 4f) (Haberman et al., 2014). In T cells, we noted that the cytotoxic genes GNLY and GZMA were amongst the most significantly upregulated in pediCD, with almost no genes downregulated (Supplemental Figure 4f).

We then systematically re-clustered each broad cell type, identifying increasing cellular heterogeneity. Given that we detected changes in the frequency of HLA-DR+ macrophages/dendritic cells and pDCs between pediCD and FGID by flow cytometry, we initially focused on the myeloid cell type sub-clustering, containing dendritic cells, macrophages, monocytes, and pDCs (Supplemental Figure 4g). Working within this analysis paradigm revealed that a traditional clustering approach had difficulty identifying the boundaries of clusters, and whether a cluster composed primarily of pediCD rather than FGID cells represented a unique cell subset, or a cell state overlaid onto a core cell subset gene expression program (Supplemental Figure 4g, Methods). These initial findings also raised the possibility that subjective analytical decisions in this manual joint clustering approach might obscure some of the unique biology of FGID and pediCD. For instance, this analytical approach could lead to hybrid clusters, informed by cells from both FGID and pediCD, that according to recent Human Cell Atlas-scale efforts in the lung, may limit discovery of rare but critical disease-specific cell types, subsets, or states (Sikkema et al., 2022).

ARBOL automated iterative tiered clustering

In order to approach the analytic challenge from a more principled direction, we made four key changes to our workflow. First, as FGID represents a non-inflamed condition and pediCD and inflamed state, we proceeded to analyze FGID and pediCD samples separately to define corresponding cell type, subset, and state clusters and markers in order to respect the underlying disease biology and maximize the potential for high-quality annotation of cell subsets. Second, to minimize analyst-to-analyst variability in the choice of clustering parameters, which can significantly under- or over-cluster data, we developed ARBOL (https://github.com/jo-m-lab/ARBOL). ARBOL is an automated clustering approach—built with conventional Seurat functions—to maximize the silhouette score at each tier of iterative sub-clustering and stop sub-clustering when a specific granularity is reached (Supplemental Figure 5a,b; Methods), Third, we systematically generated descriptive names for cell types and subsets together with differentiating marker genes. Fourth, we accounted for the number and diversity of patients which compose each cluster using Simpson’s Index of Diversity. This enabled us to focus our study on clusters that, despite being rare, were reproducibly found in the majority of patients (Simpson, 1949). Using ARBOL, each tier of analysis is typically under-clustered relative to traditional empirical analyses, but the automation proceeds through several more tiers (typically 4 to 5) until stop conditions (e.g. cell numbers and differentially expressed genes; Methods) are met. Intriguingly, this number of tiers was also manually selected in the HCA expert-annotated lung cell atlas, where reliable and robust cell clusters were found in the 0.05% cell frequency range (Sikkema et al., 2022).

We then inspected all outputs (182 FGID and 425 pediCD clusters) and provided descriptive cell cluster names independently for FGID and pediCD (Supplemental Figures 5b and 6). We also focused at this stage on flagging putative doublet clusters or clusters where the majority of differentially expressed genes which triggered further clustering consisted of known technical confounders in scRNA-seq data (e.g. mitochondrial, ribosomal, and spillover genes from cells with high secretory capacity) yielding a final number of 118 FGID and 305 pediCD clusters (Supplemental Figure 5b) (Smillie et al., 2019). Importantly, this quality-control step could often be missed by traditional scRNA-seq analysis, since these analysis paradigms retain such cells within higher-tier clusters, yet their inclusion of these cells would confound both differential composition and expression analyses if not identified. We note this clustering method represents a data-driven approach, though it may not always reflect a cellular program or transcriptional module of known biological significance or stability. Nevertheless, we highlight that many of these small clusters (∼100 cells; 0.1% of total cells) which may represent transient cell states, are found across all FGID or pediCD patients surveyed, thus suggesting a level of consistency across diseases, and across genetically and environmentally unrelated individuals.

Two comprehensive atlases of FGID and pediCD

To generate the final dendrograms for the FGID and pediCD cell atlases, we hierarchically clustered all end cell states, and performed 1 vs. rest within-Tier 1 clusters (i.e. broad cell types) differential expression to provide systematic names for cells. The naming heuristic was based on the cell type classification, cell subset classification with literature-based markers, and two genes using a rank-score function (Figures 3 and 4; Supplemental Figure 6; Methods). We provide complete gene lists at three levels of comparison: cell types (1 vs. rest across all cells), subsets (1 v. rest across all cell types), and states (1 vs. rest within-Tier 1 cell-type) in Supplemental Tables 5-10. More specifically, to select marker genes for naming in a data driven manner, we used 1 vs. rest within-cell-type differential expression (Supplemental Tables 7 and 10; Wilcoxon, Bonferroni adjusted p<0.05), and a rank-score function (-log(sig+1) * avg_logFC * (pct.in / pct.out)) to prioritize the top 2 marker genes. For example, using this ranking system, we identified CCL3 and CD160 as two genes significantly enriched in one NK cluster (adj. p-value = 0, expression within-cluster >40% cells positive and in other Tier 1 T cells <6%). This resulted in a final name for this cluster of CD.NK.CCL3.CD160. We repeated this process for all FGID and pediCD cells within Tier 1 B cell, endothelial, epithelial, fibroblast, plasma cell, myeloid cell, mast cell, and T cell identified clusters, and provide systematically generated names for all (Supplemental Figure 5; Supplemental Tables 11 and 12).

A Comprehensive Cell Atlas of Terminal Ileum in Non-Inflammatory FGID.

a. tSNE of 99,488 single-cells isolated from terminal ileal biopsies of 13 FGID patients. Colors represent major cell type groups determined via Louvain clustering with resolution set by optimized silhouette score. b. tSNE as in a with individual patients plotted. For specific proportions please see Supplemental Figure 4. c. tSNE of each major cell type which was used as input into iterative tiered clustering (ITC). d. End clusters determined by ARBOL of complete FGID data set were hierarchically clustered on the median expression of 4,445 pairwise differentially expressed genes, using complete linkage and distance calculated with Pearson correlation, between each end cell cluster. Simpson’s Index of Diversity represented as 1-Simpson’s where 1 (black) indicates equivalent richness of all patients in that cluster, and 0 (white) indicates a completely patient-specific subset. Numbers represent the number of cells in that cluster. Names of subsets are determined by Disease.CellType.GeneA.GeneB as in Methods. e. Dot plot of 2 defining genes for each cell type. Dot size represents fraction of cells expressing the gene, and color intensity represents binned count-based expression level (log(scaled UMI+1)) amongst expressing cells. All cluster defining genes are provided in Supplemental Tables 5 to 7.

A Comprehensive Cell Atlas of Terminal Ileum in pediCD.

a. tSNE of 124,054 single-cells isolated from terminal ileal biopsies of 14 pediCD patients. Colors represent major cell type groups determined via Louvain clustering with resolution set by optimized silhouette score. b. tSNE as in a with individual patients plotted. For specific proportions please see Supplemental Figure 4. c. tSNE of each major cell type which was used as input into iterative tiered clustering (ITC). d. End clusters determined by ARBOL of complete pediCD data set were hierarchically clustered on the median expression of 1,844 pairwise differentially expressed genes, using complete linkage and distance calculated with Pearson correlation, between each end cell cluster. Simpson’s Index of Diversity represented as 1-Simpson’s where 1 (black) indicates equivalent richness of all patients in that cluster, and 0 (white) indicates a completely patient-specific subset. Numbers represent the number of cells in that cluster. Names of subsets are determined by Disease.CellType.GeneA.GeneB as in Methods. e. Dot plot of 2 defining genes for each cell type. Dot size represents fraction of cells expressing the gene, and color intensity represents binned count-based expression level (log(scaled UMI+1)) amongst expressing cells. All cluster defining genes are provided in Supplemental Tables 8 to 10.

From the 99,488 cells profiled from 13 FGID patients, we recovered 12 Tier 1 clusters, representing the main cell types found in the lamina propria and remnant epithelium contained in an ileal biopsy, which we display on a t-stochastic neighbor embedding (t-SNE) plot colored by cluster identity (Figure 3a; Supplemental Figure 5b). From the 124,054 cells profiled from 14 pediCD patients, we recovered 12 Tier 1 clusters which we also display on a t-SNE plot colored by cluster identity (Figure 4a; Supplemental Figure 5b). Distinct from FGID, Paneth cells clustered separately at Tier 1 in pediCD, while glial cells were now found within the fibroblast Tier 1 cluster. Inspecting each individual patient’s contribution to each broad cell type, we noted that all patients contributed to all Tier 1 clusters in both FGID and pediCD (FGID: Figure 3b; pediCD: Figure 4b; Supplemental Figure 4c,d; Methods Supplemental Table 13). As patient identity did not factor into ARBOL stop conditions, we then calculated Simpson’s Index of Diversity for each of the end clusters (Figure 3d; Supplemental Figure 5). Although low diversity clusters may still reflect important biology for individual patients, we comment more extensively on clusters with high patient diversity. In pediCD, most end clusters are conserved across multiple patients, while only 16/305 clusters were single-patient clusters (Figure 4d; Supplemental Figure 5). We note that in pediCD, relative to FGID, a higher fraction of cell clusters exhibited lower patient diversity. In the Methods section, we provide full explanations for how names for subsets and clusters were defined for B cells, myeloid cells, T/NK/ILC’s, epithelial cells, endothelial cells, fibroblasts, mast cells and plasma cells, and include links to corresponding Supplemental Tables containing full supporting gene lists and data visualization. We share portal links in Data Availability.

Using this analytical workflow, we present two comprehensive cellular atlases, separately mapping FGID (Figure 3) and pediCD (Figure 4). The pediCD atlas enabled us to nominate cell states associated with disease severity and treatment outcomes in this disease (Figures 5 and 6), identify histological relationships between epithelial/myeloid/lymphoid cells (Figure 7), and understand how anti-TNF pushes the pediatric treatment-naïve ecosystem towards an adult tissue state (Figure 8).

A Collective Cell Vector in pediCD Reveals Predictive Axes of Disease Trajectory and Treatment Response.

a. Spearman rank correlation heatmap of principal components calculated from the frequencies of each end cluster per main cell type together with clinical metadata. Correlation is represented by both the intensity and size of the box and those which are FDR < 0.05 have a bounding box. b. A PCA plot of each patient’s end cell cluster composition as determined by ARBOL for T/NK/ILC, from the pediCD dataset where the end cell clusters frequency is calculated amongst all cell types and colors represent anti-TNF response. (inset highlights the specific correlation between PC2 of the T, Myeloid, Epithelial cell frequency analysis with anti-TNF response). We term this principal component “pediCD-TIME”. c. Cell cluster frequencies of the parent cell type found to be significant by Mann-Whitney U test between selected clusters (see Supplemental Figure 7 for all graphs; Supplemental Table 15). d. Heatmap showing cell frequencies per patient of most positive and negative cell subsets of PC2 from PCA performed on T/NK/ILC, myeloid and epithelial cell subsets (Supplemental Table 16). Cell subsets are sorted by PC2 score, and patients were sorted by anti-TNF response. Heatmap is not normalized and displaying the log counts-per million of each cell subset normalized per cell type. *Patient p022’s response category changed from FR to PR after database lock in December of 2020. No other patient’s categorization has changed.

ARBOL analysis of joint FGID and pediCD T/NK/ILCs, Epithelial and Myeloid cells identifies high-diversity clusters that drive disease severity signature.

a. A principal component analysis (PCA) plot of each patient’s end cell cluster composition as determined by ARBOL for T/NK/ILC, Myeloid, and Epithelial cells from FGID and pediCD patients where the end cell clusters frequency is calculated amongst total cells and colors represent disease state and treatment response. b. The Spearman rank correlation between PC1 of the joint T/NK/ILC, Myeloid, and Epithelial cell frequency analysis with anti-TNF response. c. A UMAP plot of 7,366 activated T/NK/ILC cells from FGID (colored by disease) and pediCD (colored by treatment response) data sets. d. Hierarchical clustering of activated T/NK/ILC cells from FGID and pediCD data set with input clusters determined based on results of ARBOL, and performed on the median expression of all genes, using complete linkage and distance calculated with Pearson correlation, between each end cell cluster. Simpson’s Index of Diversity represented as 1-Simpson’s where 1 (black) indicates equivalent richness of all patients in that cluster, and 0 (white) indicates a completely patient-specific subset. Numbers represent the number of cells in that cluster. Names of subsets are determined by Disease.CellType.GeneA.GeneB as in Methods, and names in red indicate strongest PR-PC1-negative loadings. Pie charts represent the fractional composition of FGID or pediCD cells (note the initial proportions for reference). e. Top JOINT-PC1-negative clusters are both (left) high diversity and (right) almost exclusively found in pediCD patients.

Histological and scRNA-seq analyses indicate CD4+ T cell infiltration into the epithelium of pediCD samples.

a. Representative (left) multiplex immunofluorescence of B cells (CD20), T cells (CD3), cytotoxic T cells (CD8), regulatory T cells (FOXP3), myeloid cells (CD68) and epithelium (panCK) which was then (right) analyzed using an automated classifier mask to divide sections into epithelium, lamina propria and glass areas for cellular quantification. b. Representative CD and FGID sections stained with indicated markers, inset focuses on T cells within epithelium. Quantification of 8 pediCD and 10 FGID participants is represented as a row z-scored hierarchically-clustered heatmap for total number of indicated cells per mm2 of whole section, epithelium, or lamina propria. c. Cells per mm2 of epithelium for CD3+ cells, CD3+CD8-cells (CD4 T cells), and CD3+CD8-FOXP3-(effector CD4 T cells) for pediCD and FGID participants. d. Spearman rank correlation heatmap of the counts-per-million for each of the top 25 clusters defining pediCD-TIME positive (NOA-associated) and pediCD-TIME negative (PR-associated) vectors. Correlation is represented by both the intensity and size of the box and those which are FDR < 0.05 have a bounding box.

Anti-TNF treatment shifts the pediCD cellular ecosystem towards adult treatment-refractory disease.

a. GSEA analysis showing the ranks of 92 PREDICT markers (markers of top 25 cell states associated with disease severity and treatment outcomes) in bulk RNA sequencing of illeal or colonic mucosa of two other treatment-naïve cohorts (pediatric RISK cohort, n = 69, adult E−MTAB−7604 cohort, n = 43) comparing pediCD patients who did or did not respond to anti-TNF therapy. P-value is estimated based on an adaptive multi-level split Monte-Carlo scheme. b. UMAP plots of 275,544 cells from PREDICT ileal biopsies (Baseline: 107,432 cells; Repeat: 47,796 cells) or GIMATS anti-TNF treated ileal resections (GIMATS Uninflamed: 61,965; GIMATS Inflamed 58,351) integrated using Harmony (batch = 10x version). c. A UMAP plot from cells in (b) colored by major cell type. d. A PCA plot (left) or UMAP plot (right) of each patient’s end cell cluster composition as determined by ARBOL with Harmony integration (batch = 10x version) for T, B, Myeloid, Fibroblast, Plasma from the treatment-naïve PREDICT pediCD samples (n=13), repeat PREDICT biopsies (n=2 NOA, 6 on-anti-TNF), and adult on-treatment GIMATS biopsies (n=22) where the end cell clusters frequency is calculated amongst total cells and colors represent severity and cohorts are shapes. e. A UMAP trajectory of the CPM table underlying (d). f. The top 50 varying (yellow) cell clusters along the pseudotime trajectory as determined by Moran’s I test (q-value < 0.000029). g. Cells per million of significantly varying T/NK/ILC clusters where individual points represent patient biopsies colored by severity and cohorts as shapes (as in d).

Clinical variables and cellular variance that associates with pediCD severity

As the pediCD atlas was curated from treatment-naïve diagnostic samples, we were able to interrogate the data to test if overall shifts in cellular composition, specific cell states, and/or gene expression signatures underlie clinically-appreciated disease severity and treatment decisions (NOA vs. FR/PR), and those that are further associated with response to anti-TNF therapies (FR vs. PR). Here, we leveraged the detailed clinical trajectories collected from all patients as the ultimate functional test: resolving how cellular composition and cell states anticipate disease and treatment outcomes.

In order to capture the overall principal axes of variation explaining changes in cellular composition, we calculated the fractional composition of all 305 end cell clusters in pediCD within its parent cell type (“per cell type”), or within all cells (“per total cells”), and performed a principal component analysis (PCA) over both of these sample x cell cluster frequency tables (Supplemental Table 14) (Mathew et al., 2020). We then used the PC1 (13.4% variation “per cell type” and 13.5% variation “per total cells”) and PC2 (12.7% variation “per cell type” and 11.8% variation “per total cells”) values for each sample as numerical variables which we correlated with clinical metadata (Figure 5a, r by Spearman-rank). Amongst the clinical variables, we noted strong correlation between Initial wPCDAI and CRP (r=0.83, FDR<0.05), and moderate correlation between Initial wPCDAI and either anti-TNF within 90 days (r=0.65) or anti-TNF response coded as NOA (0) FR (1) and PR (2) (r=0.49). The pediCD atlas also enabled us to correlate the transcriptome with initial disease severity and treatment response: Thus, for PC1-“per total cells”, we identified strong correlations with anti-TNF treatment within 90 days (r=-0.76), and moderate correlation anti-TNF_NOA_FR_PR (r=-0.58; Methods).

Discovery of pediCD-TIME: a collective cell vector that anticipates CD clinical severity and response spectrum

In order to understand if multiple cell types, acting in a concerted manner, were predominantly driving the association with clinical disease severity at initial presentation of pediCD, we then further deconvoluted the overall PCA on 305 end clusters and performed PCA over each cell type’s fractional composition of end clusters individually (B cells: 33 clusters, Endothelial: 18 clusters, Epithelial: 68 clusters, Fibroblast 45 clusters, Myeloid: 54 clusters, T/NK/ILCs: 57 clusters), and correlated the first two PCs (all PC1s and PC2s each accounted for >13% variance) with all of the clinical variables (Supplemental Table 14). The PCs derived from T/NK/ILC cells, myeloid cells, and epithelial cells were all moderately correlated with anti-TNF_NOA_FR_PR status (r>0.49) individually, and had higher values than the other cell types; therefore, we asked if a PCA-based metric considering all three cell types would synergistically capture both disease severity and treatment response. When we calculated the PCA accounting for frequencies of each cell subset of T/NK/ILC cells, myeloid cells, and epithelial cells amongst all cell types together, we found strong correlation for PC2 with both anti-TNF within 90 days (r=-0.83) and anti-TNF-NOA_FR_PR status (r=-0.87, FDR<0.05) (Figure 5a,b). This represented the two strongest correlations of any variable we tested with anti-TNF treatment and response status. This “PC2-T/NK/ILC/Myeloid/Epithelial” is further referred to as ‘pediCD T cell, innate lymphocyte, myeloid and epithelial’ (pediCD-TIME).

To further deconstruct pediCD-TIME, we used frequency-based statistics and analysis of PC loadings. We first identified which cell clusters accounted for the most significant changes in relative frequency based on the relative frequency of an end cell cluster within its parent cell type, noting limitations related to dissociation-induced biases (Discussion) (Gomariz et al., 2018). We performed a Fisher’s exact test between NOA vs. FR; NOA vs. PR; or FR vs. PR, and then performed a Mann-Whitney U test to highlight specific clusters (Fig. 5c; Supplemental Figure 7; Methods). As a comparison to differential expression within a cell type or subset, we highlight the power of ARBOL to deconstruct cell clusters that are often unified through strong cell-state signatures, such as proliferation, into identifiable, diverse, and disease-associated cell subsets (Supplemental Figure 8; Methods). This cluster-level analysis identified concerted changes when comparing FRs and PRs to NOAs, especially within the T/NK/ILC compartment, with PRs relative to NOAs having additional significant changes within the myeloid compartment (Fig. 5c; Supplemental Figure 7; Methods).

As noted above, higher disease severity was inversely related with the PC loadings in pediCD-TIME. Amongst the top negative PC loadings for pediCD-TIME (Fig. 5a,b), enriched in FRs—and further enriched in PRs—compared to NOAs, included both helper and cytotoxic T cell clusters (CD.T.MAF.CTLA4; CD.T.CCL20.RORA; CD.T.GNLY.CSF2), NK cell clusters (CD.NK.GNLY.FCER1G; CD.NK.GNLY.IFNG; CD.NK.GNLY.GZMB); proliferating T cells and NK cells (CD.T.MKI67.FOXP3; CD.T.MKI67.IFNG; CD.T.MKI67.IL22; CD.NK.MKI67.GZMA), and monocytes, macrophages, DCs and pDCs (CD.cDC2.CD1C.AREG; CD.Mac.C1QB.CD14; CD.Mono.CXCL3.FCN1; CD.pDC.IRF7.IL3RA; CD.Mono/Mac.CXCL10.FCN1) (Figure 5d; Supplemental Table 16). The top positive loadings for pediCD-TIME encompassing the NOA-enriched clusters included several epithelial cell subsets such as Tuft cells (CD.EC.GNAT3.TRPM5) and those with specialized metabolic features including retinol-binding, bile binding and export, fatty-acid and cholesterol metabolism, fructose and glucose metabolism, starch metabolism, glutathione metabolism, sulfation, and the terminal degradation of peptides (CD.EC.RBP2.CYP3A4; CD.EC.FABP6.PLCG2; CD.EC.FABP1.ADIRF; CD.EC.GSTA2.TMPRSS15) (Figure 5d; Supplemental Table 16) (Lampen et al., 2000; Mårtensson et al., 1990; Martínez-Augustin and de Medina, 2008; Sullivan et al., 2021; Wen and Rawls, 2020). Furthermore, clusters also enriched in NOA-pediCD-TIME such as CD.EC.ADH1C.RPS4Y1 and CD.EC.ADH1C.GSTA1, clustered in a separate branch together and expressed several enzymes responsible for steroid hormone and dopamine biosynthesis (Figure 4d, 5d) (Cima et al., 2004; Magro et al., 2002). Importantly for the regenerative potential of the epithelium, CD.EpithStem.LINC00176.RPS4Y1 were also defining of the NOA-pediCD-TIME direction.

In order to determine the relative contributions and stability of each cell cluster to the pediCD-TIME vector, we performed leave-one-out cross-validation (LOOCV) on the top-weighted pediCD-TIME clusters based on their frequency within total cells (Supplemental Figure 9a). This identified the stability of cell clusters such as CD.T.MAF.CTLA4, CD.Mono.CXCL3.FCN1, CD.T.GNLY.CSF2 and other top PR-pediCD-TIME-negative subsets. Furthermore, we also re-calculated PC weights as frequencies of clusters within their major cell type, with the standard deviation of PC loadings stabilizing even further (Supplemental Figure 9b). This suggests that compositional scRNA-seq analysis provides more robust signatures within major cell types than amongst all total cells, and both methods can recover a disease trajectory.

Together, these discoveries underscore the multiple collective changes in the composition and/or state of T/NK/ILC cells, myeloid cells, and epithelial cells at pediCD diagnosis that could stratify pediCD patients by disease severity, and may influence anti-TNF responsiveness.

Cluster-based and cluster-free approaches contextualize pediCD with FGID

As we had generated independent cellular atlases for FGID and pediCD to mitigate aspects of disease batch effects obscuring the ability to resolve cell clusters, we next sought to compare pediCD and FGID cell subsets using label-retaining (Random Forest, ARBOL) and label-free (diffusion maps) methods. We applied these models to each cell type individually, and here focus our discussion on Myeloid cells and T/NK/ILC cells as two cell types prominently associated with pediCD disease severity (Supplemental Figures 10-12). As newer methods are developed, more refined integration is likely to be possible. Comparing across myeloid cells between pediCD and FGID, we could identify strong correspondence of specific cell subsets such as cDC1s or pDCs (Supplemental Figure 10a). We also identified strong correspondence between several cDC2 clusters. We identified a gradient of monocyte and macrophage correspondence of 31 clusters in pediCD to 2 FGID clusters, likely reflective of inflammatory monocyte to macrophage differentiation in pediCD (Blériot et al., 2020; Dutertre et al., 2019; Guilliams et al., 2018). For T/NK/ILC cells, we identified more discrete patterns relative to Myeloid cells based on comparison of the RF result. Within the two FGID cytotoxic T cell clusters, we identified correspondence by 18 pediCD clusters, representing ILC3s, and cytotoxic NK cells and T cells (Supplemental Figure 11). The cluster of naïve T cells in FGID had correspondence with the majority of pediCD non-cytotoxic T cell clusters, illustrating a substantial activation and specialization to several discrete T cell states that were specific for pediCD.

We next performed an analysis over a shared gene expression space of FGID and pediCD of the monocytes/macrophages (Supplemental Figure 13) and T/NK/ILCs (Supplemental Figure 14). Within FGID monocytes/macrophages, we identified that the majority of clusters occupied the periphery of the UMAP space, including chemokine-expressing clusters (FG.Mac.CCL3.HES1; FG.Mac.CXCL8.IL1B) and metabolic clusters (FG.Mac.APOE.PTGDS) (Supplemental Figure 13a,c). This was in stark contrast to the pediCD monocytes/macrophages, where we identified that now many of the clusters occupied the central region of the UMAP (Supplemental Figure 13a,c). We highlight several of the clusters that are significantly different in frequency between the pediCD groups which were found in this central region (Supplemental Figure 13b), and how NOA, FR and PR pediCD clusters had significantly different distributions within this space as measured by the Hellinger Distance (Supplemental Figure 13c-e; Methods). Of note, the frequency of TNF+ macrophages and its expression level of TNF was significantly increased in FRs relative to all other groups (Supplemental Figure 13f; permutation test shuffling anti-TNF response variable, FR had significantly more TNF+ cells than expected by chance with p approximating 0). Despite the fine-grained tiered clustering approach used, the majority of clusters had high Simpson’s Diversity indices representing cell states found in several patients (Supplemental Figure 13g).

Within T/NK/ILCs, we identified that FGID cells were more uniformly mixed with the pediCD cells relative to monocytes/macrophages (Supplemental Figure 14a). FGID cells occupied naïve and quiescent states, showed some signatures of activation, and also specialization towards helper and cytotoxic states (Supplemental Figure 14a,c) (Sallusto et al., 1999). The most notable changes in the Hellinger Distance distribution occurred between FGID and FR, rather than between FGID and PR (Supplemental Figure 14d,e). Similar to the monocytes/macrophages, the main area which gained density with increased disease severity was the central region: characterized by proliferation of several clusters increased in frequency within FRs and PRs, including T cells such as CD.T.MKI67.FOXP3 and CD.T.MKI67.IL22 (Supplemental Figure 14b-d). Taken together with several cell clusters associated with pediCD proliferation overlapping with existing areas found in FGID, this indicates activation and more extreme diversification from existing T cell states driving the T cell clustering that defined pediCD. This is distinct from the recruitment and failed differentiation towards homeostatic cell states between FGID and pediCD that we discovered in monocytes/macrophage clusters. Both joint projections confirm and extend our random forest correspondence probabilities, and provide a clustering-independent view of how cellular density is shifted between FGID and pediCD (Dann et al., 2022).

Joint ARBOL atlas of activated T/NK/ILC’s in FGID and pediCD

Building on our Random Forest and Hellinger Distance distribution analyses, we sought to formally cluster pediCD and FGID T/NK/ILC, Myeloid, and Epithelial cells together to place the pediCD patients within context of FGID, taking advantage of the high-resolution view afforded by ARBOL (Fig. 6a). We tested whether performing ARBOL on a joint FGID and pediCD dataset would also discern a disease severity vector. We identified that JOINT-PC1 in this analysis captures a disease severity vector, with the positive direction enriched in FGID patients, and the negative direction enriched in FR and PR pediCD patients (Fig. 6b; r= −0.74, p<0.0001). We focused on a branch of 7,366 activated T/NK/ILCs, annotated all end clusters as T, NK or ILC (Supplemental Figure 6), and selected 2 genes from the top-10 rank-scored list (Methods) that provide a descriptive name for each cluster (Fig. 6c,d). We visualized at each split of the binary tree representation the fractional composition of cells weighted towards FGID or pediCD cells.

Our tree recovered primary splits of memory T cells, cytotoxic cells, helper cells, and intraepithelial cells, with proliferative cells interspersed within these groups. Intriguingly, this analysis reinforces our flow cytometric and initial two-tiered manual clustering approach whereby up to a certain level of the tree, FGID and pediCD do not show significant differences in composition. However, when following the additional splits leading to each end cluster, we now note several clusters which are pediCD or FGID enriched. Furthermore, despite this disease-enrichment, the end clusters are still high in patient diversity, with the majority falling above 0.5 Simpson’s Index (Figure 6d,e). Importantly, our top-ranked PR-PC1-negative T cells contributing to the disease severity vector are all high diversity clusters even when clustered jointly with FGID (Fig. 6; Supplemental Figure 15). We again recover a majority of proliferating T and NK cell clusters contributing to the JOINT-PC1 disease severity vector (NK.MKI67.CD38, T.MKI67.FABP5, T.MKI67.IL17A, T.MKI67.IL26, T.MKI67.FOXP3, and others).

The joint FGID-pediCD atlas showed an overlap between less severe pediCD patients (NOA) and FGID samples. To understand what causes the alignment between these groups, we used an automated naming method to name the 802 clusters in the joint ARBOL, and plotted the subsets with the most positive loadings in JOINT-PC1 (Supplemental Figure 15). We used alluvial diagrams to match the joint cell states with their corresponding curated names from the CD atlas to contextualize the FGID samples in terms of the NOA samples. Doing so revealed subsets of CCR7+ T cells and CLEC10A+ cDC2’s are shared between these groups that are less prominent in severe CD (Supplemental Figure 15e). However, in order to determine if these cell states are distinct from healthy T and DC states, we would need to directly integrate healthy data.

While further experiments would need to be conducted to isolate and determine the stable or transient nature of each cell cluster, their conservation across multiple patients from distinct genetic and environmental backgrounds suggests that these end clusters may be reliably found across most patients sampled within a disease category, and that the primary disease-associated signatures remain robust when placed in context of another pediatric gastrointestinal disease. As we had now generated and annotated three atlases: FGID, pediCD, and joint FGID/pediCD, we took the opportunity to calculate how considering atlases separately or jointly influenced the coherence of each end cluster (Supplemental Figure 15; Methods). We anticipate that as atlases scale in cell number, patient number, cohort number, and meta-analyses, that multiple methods of clustering will be important to fully-resolve the cellular heterogeneity present within each sample, while retaining community-level labels for facile integration and comparison for meta-analyses.

Pair-Wise Correlation of Cell Clusters and Histology

In order to provide spatial quantification for key cell types in our disease signature, we also utilized multiplex immunofluorescence to identify the absolute numbers and locations of B cells, cytotoxic, helper and regulatory T cells, and macrophages in FGID and pediCD samples. Using an automated mask for epithelial and lamina propria identification, we were able to resolve the location of these cell types within the whole section or within the epithelial or lamina propria layers of the tissue (Figure 7a). We identified no significant differences in the whole sections for total numbers of T cells (CD3+), B cells (CD20+), myeloid cells (CD68+), and cytotoxic T cells or regulatory T cells between pediCD (n=8) and FGID (n=10) participants (Figure 7b). Strikingly, we did find a significant difference in total CD3+ cells within the epithelium, and specifically of CD3+CD8-FOXP3-(CD4+ effector T cells) using our multiplex immunofluorescence panel (Figure 7c). This indicates that CD4+ cells infiltrating into the epithelial layer may have an early and outsized role in driving pediCD.

In order to understand the impact of which CD4+ T cell states that were associated with anti-TNF response may be influencing specific epithelial states, we correlated proliferating T and NK cell clusters with epithelial cell clusters (Figure 7d). We used the single-cell data for this, where we are powered to make comparisons between response groups. We found that CD.T.MKI67.FOXP3 were strongly positively associated with CD.Goblet.RETNLB.ITLN1 and CD.EC.NUPR1.LCN2 secretory cell states. Conversely, we found that CD.T.MKI67.IL22 were significantly negatively correlated with CD.EC.GNAT3.TRPM5 and EC.FABP1.ADIRF cells, and that CD.T.CCL20.RORA were significantly negatively correlated with EC.ADH1C.GSTA1 and EpithStem.LINC00176.RPS4Y1. This indicates the potential for CD4 T cells present within the epithelial layer to significantly alter epithelial cell states away from nutrient sensing and homeostatic metabolic function in pediCD.

Contextualization to published Crohn’s Disease Studies

To determine whether the pediCD-TIME disease severity gene signature that we discovered in the PREDICT study can be found in other cohorts with published bulk RNA-seq data, we selected the top 92 markers of the 25 cell states associated with disease severity and treatment outcomes (Methods; Supplemental Table 17) and performed a gene-set enrichment analysis (GSEA) (Figure 8a; Supplemental Figure 16). We used the cell clusters ranked by cell frequency of total cells, as this would be most representative of bulk-analyzed tissue. In the PREDICT cohort, we were able to validate the 92-gene signature as significantly enriched within FR and PR patients relative to FGID or NOA (all p<9.3×10-05). These 92 genes were also enriched in PR relative to FR patients (p<4.3×10-05). Notably, in the two independent treatment-naïve cohorts that we analyzed (the pediatric RISK cohort, n = 69, and the adult E−MTAB−7604 cohort, n = 43) this 92-gene signature was significantly enriched in illeal, but not colonic, mucosal biopsies from patients who did not respond to anti-TNF therapy compared to those who responded (Figure 8a; p= 0.0046 vs. p=0.78, respectively) (Kugathasan et al., 2017; Verstockt et al., 2019). Thus, these 92 genes (including TNFAIP6, GZMB, S100A8, CSF2, CLEC4E, S100A9, IL1RN, FCGR1A, CLIC3, CD14, PLA2G7, FAM26F, IL3RA, NKG7, IL32, CCL3, OLR1, LILRA4, APOC1, MYBL2 and others; Supplemental Table 17) nominate a signature of anticipatory markers of anti-TNF therapy outcome in newly-diagnosed patients and are validated in one internal (bulk RNA-seq from PREDICT patients) and two external (RISK and E-MTAB-7604) bulk RNA-seq cohorts.

Anti-TNF treated repeat pediCD biopsies are closer to adult treatment-refractory CD

We sought to test how anti-TNF treatment shifts the cellular ecosystems of the pediatric ileum. The availability of repeat biopsy specimens from PREDICT patients enabled us to perform a joint ARBOL containing treatment-naïve pediCD biopsies (n=14), 8 follow-up endoscopy samples (from 2 NOA and 6 on-anti-TNF PREDICT patients), and 11 adult CD patients from the (Martin et al., 2019) study which identified the GIMATS (T, B, Myeloid, Fibroblast Plasma) cellular module (Figure 8b).

Due to the utilization of several 10X 3’ kit versions, we added the integration method Harmony into the ARBOL pipeline (Figure 8b,c; Methods). Of note, pediCD atlas curated clusters were preserved with a similar distribution of Shannon entropy in the integrated atlas compared to the joint atlas (Supplemental Figure 15). This analysis illustrated that we continue to recover a gradient of disease severity based on our patients’ cell clusters from NOA, to untreated FR and PR. Intriguingly, the repeat anti-TNF treated samples begin to occupy the area between the baseline PREDICT samples and the on-treatment GIMATS adult samples (Figure 8d; diamonds are repeat samples). Meanwhile, the two NOA patients with repeat biopsies remain at the beginning of the trajectory overlaid next to their corresponding baseline samples.

From the resultant integrated ARBOL, we calculated a pseudotime trajectory of patients informed by cell clusters. By calculating the cell clusters most auto-correlated along the pseudotime trajectory, this revealed that T/NK/ILC, Myeloid, and Epithelial cell clusters (i.e. the same cell types defining the pediCD-TIME signature) were enriched in those tracking with the trajectory (Figure 8e,f). By using the automated naming function within ARBOL, we were able to provide identities for these cell clusters, and present those for T/NK/ILC cells (Figure 8g). Our analysis revealed that T.KLF2.FAM65B.CCR7.SELL (expressing markers of naïve T cells), T.GZMK.CCL5.HLA-B.TRAC (expressing markers of CD8 cytotoxic T cells) and T.S100A4.CCR6.TMSB4X.PFN1 (expressing markers of Th17 cells) were significantly decreased with anti-TNF treatment in pediatric and adult patients. Correspondingly, T.IL7R.TNFAIP3,KLRB1 (with signatures of TNF activation), T.CCL5.CD8A.CD8B.TXNIP (effector CD8 T cells), and T.KIT.PCDH9.LINC00299.AREG; T.LST1.AREG.CXCL2.LINC00299 (innate lymphoid cell clusters) increased.

To further characterize which cell states exhibited consistent trends across this disease-associated pseudotime, we conducted additional validation analysis using the TradeSeq package (Supplemental Figure 17). Through leave-one-out cross-validation and we assessed the robustness of the association between each cell state and the pseudotime vector. TradeSeq’s trajectory clustering method enabled us to classify the robustly associated cell states into distinct pseudo-temporal patterns (e.g. downward-hill, downward-slope, quick-climb, and upward-hill; Supplemental Figure 17b). Notably, bootstrap analysis confirmed that subsets of T, Myeloid, and Epithelial cells consistently exhibited strong associations with pseudotime. This analysis identifies those clusters within the terminal ileum which are most associated with compensation to TNF inhibition, and illustrates that while our NOA patients exhibited remarkable consistency in the pseudotime location of their biopsies within this trajectory, anti-TNF treatment consistently “pushed” a patient’s cells closer to the location of the anti-TNF refractory disease observed in adult CD.

Discussion

This study addresses a critical unmet need in the fields of IBD and systems immunology: the creation of an atlas of newly-diagnosed untreated diseased tissue, coupled with detailed clinical follow-up to link diagnostic cell types and states with disease trajectory. This is especially true for GI autoimmune disease, and other diseases which affect tissues not easily accessible without operative or endoscopic intervention, and where tissue-specific immune pathology dictates disease severity and trajectory. Likewise, cross-sectional studies, as have been the norm for most previous scRNA-seq studies of IBD, are not able to overlay disease trajectory and treatment response onto the topography of a complex multi-cellular atlas, thus limiting the mechanistic and predictive inferences that can be drawn from the generated atlas (Corridoni et al., 2020b, 2020a; Drokhlyansky et al., 2019; Elmentaite et al., 2020; Huang et al., 2019; Kinchen et al., 2018; Martin et al., 2019; Parikh et al., 2019; Smillie et al., 2019; Uzzan et al., 2022).

Furthermore, mouse models of CD, and of IBD more broadly, may not be the most appropriate models for understanding treatment resistance in pediCD (Neurath, 2019). To surmount these limitations, we created a prospective clinical study, and enrolled patients requiring a diagnostic biopsy for possible IBD, prior to diagnosis. This allowed us to capture a tremendously valuable control group: those patients with FGID, who experience GI symptoms, but for whom the endoscopy does not reveal evidence of GI inflammation or autoimmunity. These uninflamed controls served as a critical comparator to contextualize the evidence of immune pathology that we observed in patients with pediCD across the severity spectrum. With these detailed clinical phenotypes as our foundation, we developed an automated iterative tiered clustering algorithm for scRNA-seq data, ARBOL, which defined pediCD-TIME, a vector of T cells, myeloid cells and epithelial cells that stratifies both pediCD disease severity and response to treatment. We also relate our treatment-naïve samples to repeat samples from PREDICT patients on anti-TNF, and to an adult on-treatment Crohn’s scRNA-seq atlas, identifying a continuum between pediatric and adult Crohn’s disease that may be bridged by the effects of treatment (Martin et al., 2019).

The availability of comprehensive clinical, flow cytometric and scRNA-seq data from patients with pediCD and from uninflamed FGID controls created an unprecedented opportunity for comparative atlas creation. We developed a methodical, unbiased, approach to cell state discovery, ARBOL, released alongside our manuscript in both R (https://github.com/jo-m-lab/ARBOL) and python (https://github.com/jo-m-lab/ARBOLpy). ARBOL iteratively explores axes of variation in scRNA-seq data by clustering and subclustering until specific stop conditions are met. The philosophy of ARBOL is that multiple axes of variation could be biologically meaningful at each tier, and that axes of variation are relative to the comparative outgroup, meaning that similar cell states may arise at distinct tiers. Once these possibilities are explored, curation and a statistical interrogation of resolution are used to collapse clusters into the elemental transcriptomes and co-varying gene expression of the dataset. ARBOL inherently builds a tree of subclustering events. As data is separated by major axes of variation in each subset, later rounds capture less pronounced variables. This comes with some caveats: variation shared by all cell types (for example, cell cycle stage) can make up one of the major axes of variation in the first round of clustering. Cell types can split up at the beginning, so the same splitting of B and T cells, for example, may happen further down in separate branches. The resulting tree of clustering events (Supplemental Figure 5a) is therefore neither indicative of true distances between end clusters nor a tree of unique groupings. We address this problem by calculation of a binary tree of manually and computationally curated end clusters. Using a standardized method of end cluster naming, which we describe in ARBOL’s tutorial (https://jo-m-lab.github.io/ARBOL/ARBOLtutorial.html), we found the resulting binary tree assorted end clusters into appreciated cell types and subsets (Figures 3 and 4), and also reveals further previously unappreciated granularity that will serve as the foundation for future work into the cellular composition of the gastrointestinal tract.

One of the primary remaining challenges going forward will be to identify which clusters are truly patient-unique, or are simply patient-unique at the cohort size to which we are currently limited to. We calculate a diversity metric for each end cluster to highlight those which are largely conserved between patients, and provide complete cluster-defining gene lists for both FGID and pediCD at three levels of clustering. We also provide links to our data visualization portal to enable cross-atlas comparisons: https://singlecell.broadinstitute.org/single_cell/study/SCP1422/predict-2021-paper-fgid and https://singlecell.broadinstitute.org/single_cell/study/SCP1423/predict-2021-paper-cd.

ARBOL can be deployed in the generation of cell atlases from multiple study designs to yield, for example, separate atlases for distinct diseases, joint atlases of diseases within a study, or integrated larger meta-atlases that merge multiple studies. Recent work from the integrated Human Lung Cell Atlas highlights that under-clustering of cells continues to be a large hurdle in many individual studies, but that higher clustering power must be balanced with the use of more stringent integration methods to merge control and disease groups without obscuring known subsets within lymphoid cells such as regulatory T cells and ILCs (Sikkema et al., 2022). One of the chief advantages of enrolling patients at diagnosis, and prior to any therapeutic intervention, was that we were able to relate their diagnostic immune landscape with disease trajectory. In the pediCD group, we identified 3 clinical subgroups. The first distinction was made by treating physicians, and classified patients with milder versus more severe clinical disease characteristics at diagnosis. The milder patients were not placed on anti-TNF agents (NOA), while the more severe patients were treated with monoclonal antibodies that neutralize TNF including infliximab and adalimumab. The second distinction between patient groups could not be made at diagnosis, but rather, was based on clinical and biochemical response to anti-TNF agents. Thus, of those patients treated with anti-TNF therapeutics, some were FRs, and some were only PRs, with PRs requiring anti-TNF dose modifications and the addition of other agents, and with ongoing, uncontrolled disease signs and symptoms. While differences in anti-TNF pharmacokinetics have been partially implicated in the need to dose-escalate anti-TNF agents in some pediCD patients, our study identifies foundational differences in the immune state at diagnosis in PR patients compared to the NOA and FR subgroups (D’Haens and Deventer, 2021; Ordás et al., 2012; Yarur et al., 2016). Although standard flow cytometry was not able to distinguish the immune phenotype of NOA versus treated patients, scRNA-seq identified significant differences. The contextualization of our scRNA-seq derived predictive cellular vector, pediCD-TIME, with two other treatment-naïve bulk RNA-seq studies of Crohn’s disease, and a reference adult scRNA-seq study, underscores the broader applicability of our findings (Kugathasan et al., 2017; Verstockt et al., 2019).

We noted significant cell state changes at diagnosis underlying clinically-appreciated disease severity that impacted the clinical decision to treat or not to treat with anti-TNF agents. These occurred within multiple clusters of T, NK, fibroblast, epithelial, monocyte, macrophage, and dendritic cells. For anti-TNF response, very few clusters exhibited significantly differential composition between FR and PR individuals. This suggests that multiple collective changes in several cell types may conspire to lead to differences in treatment outcomes. Indeed, when we jointly considered a cellular principal component vector comprising epithelial cells, T/NK/ILCs, and myeloid cells, we identified several clusters that together could delineate the full spectrum of NOA, FR, and PR. This cellular vector (pediCD-TIME) indicated that multiple T cell subsets, NK cells, monocytes, macrophages, and epithelial cells were altered in disease.

Epithelial cells involved in chemosensation (Tuft.GNAT3.TRPM5) and absorption of metabolites (EC.GSTA3.TMPRSS15), as well as stem cells (Banerjee et al., 2020; Sido et al., 1998; von Moltke et al., 2016) were enriched in NOA individuals. That pediCD severity is not uniquely predicted by a singular cell subset or gene is reflective of the complex genetics and environmental factors that have been implicated, along with the rich literature that has found significant changes by histology, flow cytometry, or mass cytometry in CD relative to control tissue (Buisine et al., 2001; Leeb et al., 2003; Leonard et al., 1995; Lilja et al., 2000; Mitsialis et al., 2020; Müller et al., 1998; Souza et al., 1999; Stappenbeck and McGovern, 2017; Takayama et al., 2010). However, with the PREDICT study, we have discovered precisely which changes in CD cellular composition collectively form a vector that anticipates both disease severity and treatment response. Intriguingly, the quantification and visualization of this response vector predicted a later escalation of one of our patients (p022; who appeared as an outlier FR in Figure 5d) from FR to PR, which occurred after our database lock. Furthermore, our approach also leverages co-variation of genes across single cells to identify gene sets which are not found by differential expression testing (which ignores associations between genes across cells) yet give rise to recurrent cell states found in multiple patients that are differentially abundant.

When considering the relationships between T cells and NK cells along with epithelial cells, we captured that proliferating cytotoxic NK cell subsets like CD.NK.MKI67.GZMA were significantly negatively correlated with critical metabolic and progenitor epithelial cell subsets in pediCD. Conversely, proliferating regulatory CD.T.MKI67.FOXP3 were positively associated with secretory epithelial cells in pediCD, but did not appear related to the decrease in metabolic or progenitor cells. In addition, we have found that pediCD samples have a significant increase in the number of CD4 T cells localized to the epithelial lining relative to FGID. How T cell-derived cytokines impact intestinal regeneration and differentiation has recently been the focus of several studies, but the relationship of these fine-grained T cell subsets with specific epithelial cell states observed in the human intestine remained unknown (Biton et al., 2018; Lindemans et al., 2015). Furthermore, our study builds on work from other groups who have recently started to describe specific features of intestinal intraepithelial T cells in inflammation and infection (Hoytema van Konijnenburg et al., 2017; Jaeger et al., 2021; Parsa et al., 2022). Our work suggests that there is further complexity to understand, particularly as it pertains to specific subsets of cytotoxic NK cells and T cells, and their impact on epithelial cell homeostasis and regeneration in pediCD.

The mapping of these disease severity-associated cell networks identifies a host of new potential therapeutic targets for pediCD, for many of which there are clinical-stage therapeutics that could be investigated. These include CD40L-blocking antibodies, IL-22 agonists, and targeted anti-proliferation agents (Betts et al., 2017; Furlan et al., 2015; Lindemans et al., 2015; Miura et al., 2021; Ramanujam et al., 2020; Sootome et al., 2020). A case can also be built for targeting inflammatory cytokines such as IL-1β, and for interrogating agents aimed at mucosal healing including new anti-GM-CSF antibodies, given that several prominent cell subsets marked by CSF2 were enriched in the PR patients (Ai et al., 2021; Aschenbrenner et al., 2021; Castro-Dopico et al., 2020; Mehta et al., 2020; Mitsialis et al., 2020; Muro and Mrowiec, 2015). This atlas therefore provides a rigorous evidence-based rationale for proposing new therapeutic interventions, as well as a mechanism for interrogating the impact of new agents on the longitudinal immune landscape of pediCD patients. It also identifies those clusters which respond to anti-TNF treatment and may present new therapeutic targets.

Our analysis underscores the power in the disease-specific clustering employed with this dataset, which revealed principled end clusters in pediCD which could then be related back to a non-inflammatory reference atlas of FGID, and contextualized with published adult CD atlases; while still maintaining fine granularity through distinct computational methods. Recent work on COVID-19 has also highlighted the challenges faced by systems approaches to capture baseline cell states that predict disease trajectory (Kaczorowski et al., 2017; Lucas et al., 2020; Mathew et al., 2020; Schulte-Schrepping et al., 2020; Su et al., 2020). In a disease of known infectious etiology with SARS-CoV-2, monocytes, macrophages, granulocytes, T cells, B cells, antibodies, and interferon state have all independently been associated with disease outcomes. Enabled by larger numbers of participant biopsies profiled at single-cell resolution, recent work in cancer builds on analytical frameworks using dimensional reduction techniques on flow cytometry data to understand collective cellular changes in multi-cellular tissues (Combes et al., 2022; Kaczorowski et al., 2017; Nieto et al., 2021). This work has helped establish the paradigm of tissue “archetypes” or “immunotypes” that represent the collective cellular system. Here, we have built on this foundation to consider how multiple collective changes at baseline may influence outcome, yet are likely more reflective of the disease. With the complex and protracted presentation of a multifactorial disease like Crohn’s disease, our data suggest that multiple concerted effects are required to dictate both the severity (NOA vs FR/PR) and the treatment-response (FR vs PR). Future work will further consider which cell subsets are recovered during mucosal healing, and how closely the treated state reflects each individual patient’s baseline presentation. Our work taken together with our colleagues’ points to the emergent opportunity to place a patient’s tissue in the broader cellular context of health and disease.

Limitations of the Study

Our study has several limitations that are important to consider. Despite having profiled more treatment-naïve patients than any other studies at single-cell resolution, we are still limited by the cohort size to the strict inclusion/exclusion criteria used, follow-up period, and the ability to capture untreated patients early during the diagnosis process. As a result, we are underpowered to directly correlate in our cohort genetic and environmental influences such as the enteric microbiome, infections, and dietary factors (Dovrolis et al., 2020; Rajca et al., 2014; Yilmaz et al., 2019). Second, we are limited in our ability to capture full spatial transcriptomic and multiplexed immunofluorescence information from our initial cohort. It will also be critical to understand how disease progresses in larger numbers of patients from our cohort, and the reason for persistent NOA, FR, or PR patient states in some individuals which will require retention in the trial and additional repeat biopsy procurement from larger numbers of participants. Third, our cohort is representative of the demographics of FGID and IBD in Seattle, WA. It will be essential to understand how both FGID and IBD may present differently across the world, and multiple cohorts will be required to understand location-unique and generalizable findings. Fourth, we also recognize the number of pediCD patients within this cohort limits our ability to subdivide CD phenotype into location and behavior of disease (via the Paris/Montreal classification systems) (Levine et al., 2011). Fifth, we also highlight the rapidly-evolving field of approaches to integrate single-cell datasets towards a larger gut cell atlas and the trade-offs in resolution that occur as additional datasets are brought in that require more stringent integration methods. We recognize that integration methods, as well as the use of UMAP to represent multiple PC’s in two dimensions, which, while helpful to jointly visualize studies together, may lower the resolution, and in some cases even make cell subsets like T, NK, or ILCs challenging to discern (Sikkema et al., 2022). Future work will need to understand specific platform differences that drive integration to help specifically separate batch from biology. Finally, as we are working with small pediatric IBD biopsies, our ability to functionally test and validate the stability of these end cell clusters was limited in this study. Based on our study, we will seek to expand our efforts to validate and extend our finding in ongoing work.